---
title: "Pipeline 1: Bulk RNA-seq Analysis of CLL Dasatinib Response"
subtitle: "SummarizedExperiment + edgeR/limma-voom Approach"
author:
- name: "Samuel Escudero, Ivon Sanchez, Karim Hamed"
  affiliation: Universitat Pompeu Fabra
  email: samuel.escudero01@estudiant.upf.edu
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_captions: yes
    code_folding: show
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Pipeline1_edgeR_limma}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "",
  fig.align   = "center",
  fig.width   = 8,
  fig.height  = 6,
  fig.wide    = TRUE,
  cache       = FALSE,
  message     = FALSE,
  warning     = FALSE
)
```

# Introduction

## Biological context

Chronic lymphocytic leukemia (CLL) is a clinically heterogeneous
hematological malignancy characterized by diverse responses to targeted
treatments. Dasatinib, a second-generation tyrosine kinase inhibitor
(TKI), represents a potential therapeutic strategy for patients with CLL.
However, its clinical efficacy varies significantly between individuals.
Blatte et al. (2021) employed RNA sequencing (RNA-seq) to characterize
the transcriptional profiles of primary leukemia cells from
treatment-naive CLL patients classified into two groups based on in vitro
cytotoxicity assays: **responders** (dasatinib-sensitive) and
**nonresponders** (dasatinib-resistant). Their primary goal was to
discover gene expression signatures predictive of dasatinib sensitivity.

The RNAseq data generated by Blatte and colleagues were deposited at the
Gene Expression Omnibus (GEO), publicly accessible under accession
[GSE151159](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151159).

## Analytical approach: Why SummarizedExperiment + edgeR/limma-voom?

In this pipeline, we use the **SummarizedExperiment** container as our
central data structure, combined with the **edgeR/limma-voom** framework
for differential expression (DE) analysis. This is the approach taught in
the IEO course at UPF and represents one of the most widely used
frameworks in bulk RNA-seq analysis.

**Theory: The SummarizedExperiment object**

The `SummarizedExperiment` class from Bioconductor is designed to store
rectangular matrices of experimental results (e.g., count matrices from
RNA-seq). It keeps three critical pieces of information tightly coupled:

- `assays()` - one or more matrices (e.g., raw counts, logCPM, normalized
  values), where rows are genomic features (genes) and columns are samples
- `rowData()` - a DataFrame of gene-level annotations (Ensembl ID,
  biotype, gene symbol)
- `colData()` - a DataFrame of sample-level metadata (treatment condition,
  batch, clinical covariates)

This coupling ensures that whenever we subset genes or samples, all three
components stay in sync, preventing the misalignment errors that plague
manual matrix manipulation.

**Theory: Why edgeR + limma-voom instead of DESeq2?**

Both frameworks are statistically sound, but they model the data
differently. DESeq2 models raw counts directly using a negative binomial
distribution, while the edgeR/limma-voom approach first transforms
counts into log-CPM values with precision weights (via `voom()`), then
applies the powerful linear modeling framework of limma. The voom
transformation estimates the mean-variance relationship and generates
observation-level weights that stabilize the variance, allowing us to
use standard linear models. This approach is particularly effective when
we have moderate-to-large sample sizes and want to fit complex designs
(e.g., paired samples, factorial designs, blocking). Benchmarking
studies have shown that DESeq2 and limma-voom display comparable
performance, with limma-voom being computationally faster and more
flexible for complex experimental designs.

## Libraries

```{r load-libraries}
# --- Core Bioconductor packages ---
library(SummarizedExperiment) # Central data container
library(edgeR)                # DGEList, calcNormFactors, cpm, filterByExpr
library(limma)                # voom, lmFit, eBayes, topTable, makeContrasts

# --- Batch effect detection and correction ---
library(sva)                  # Surrogate Variable Analysis

# --- Functional enrichment ---
library(GOstats)              # Hypergeometric GO testing
library(org.Hs.eg.db)         # Human gene annotation database
library(AnnotationDbi)        # Interface to annotation resources
library(fgsea)                # Fast Gene Set Enrichment Analysis

# --- Visualization and reporting ---
library(pheatmap)             # Heatmaps with clustering
library(RColorBrewer)         # Color palettes
library(ggplot2)              # Grammar of graphics
library(ggrepel)              # Non-overlapping text labels
library(dplyr)                # Data manipulation
library(tibble)               # rownames_to_column()
library(knitr)                # kable()
library(kableExtra)           # Table styling
```

# Data Import and Initial Inspection

## Loading the SummarizedExperiment

We load the `GSE151159.rds` file, which contains the complete dataset
packaged as a `SummarizedExperiment` object. This object was assembled
from GEO and includes raw gene-level counts, sample metadata, and gene
annotations.

```{r load-data}
# Load the SummarizedExperiment
# Adjust the path depending on your setup:
# Option A: from the IEOproject R package
# cll <- readRDS(file.path(system.file("extdata", package = "IEOproject"), "GSE151159.rds"))
# Option B: from the data/ directory (recommended for this repository)
cll <- readRDS("../data/GSE151159.rds")

# Inspect the object
cll
```

**What we see**: The object contains `r nrow(cll)` genes across
`r ncol(cll)` samples, with the count matrix stored in `assays(cll)$counts`.

```{r inspect-structure}
# Examine the structure
cat("Assays available:", names(assays(cll)), "\n")
cat("Dimensions:", nrow(cll), "genes x", ncol(cll), "samples\n\n")

# Peek at the count matrix
cat("First 5 genes x 3 samples of the count matrix:\n")
assays(cll)$counts[1:5, 1:3]
```

```{r inspect-coldata}
# Sample metadata
cat("Sample-level metadata columns:\n")
cat(colnames(colData(cll)), sep = "\n")
```

```{r inspect-rowdata}
# Gene annotations
cat("\nGene-level annotation columns:\n")
cat(colnames(rowData(cll)), sep = "\n")
cat("\nFirst 5 gene annotations:\n")
head(as.data.frame(rowData(cll)), 5)
```

## Sample Classification: Defining Responders vs. Non-Responders

The original metadata uses `description.2` to encode treatment response.
We create a clean binary factor `response2` with levels
`NonResponder` and `Responder`, and remove ambiguous samples labeled "all".

```{r classify-response}
# Examine the raw descriptions
cat("Raw sample descriptions:\n")
table(colData(cll)$description.2)

# Create clean response classification
desc2 <- colData(cll)$description.2
response_raw <- ifelse(
  grepl("^responders_final", desc2), "Responder",
  ifelse(grepl("^nonresponders_final", desc2), "NonResponder", NA)
)

colData(cll)$response2 <- factor(
  response_raw,
  levels = c("NonResponder", "Responder")
)

cat("\nClean response classification:\n")
table(colData(cll)$response2, useNA = "ifany")

# Remove samples with undefined response (labeled "all")
keep_samples <- !is.na(colData(cll)$response2)
cll <- cll[, keep_samples]

cat("\nAfter removing ambiguous samples:", ncol(cll), "samples retained\n")
cat("  NonResponders:", sum(cll$response2 == "NonResponder"), "\n")
cat("  Responders:   ", sum(cll$response2 == "Responder"), "\n")
```

## Export Shared Data (for Pipelines 2 and 3)

We export the count matrix and metadata as CSV files so that the Seurat
(Pipeline 2) and Scanpy (Pipeline 3) approaches can consume the same
preprocessed data without needing the `SummarizedExperiment` package.

```{r export-csvs}
# Export counts
write.csv(as.data.frame(assays(cll)$counts),
          "../data/counts_matrix.csv", row.names = TRUE)

# Export sample metadata
write.csv(as.data.frame(colData(cll)),
          "../data/sample_metadata.csv", row.names = TRUE)

# Export gene metadata
write.csv(as.data.frame(rowData(cll)),
          "../data/gene_metadata.csv", row.names = TRUE)

cat("Exported shared CSVs to data/ for Pipelines 2 and 3\n")
```

# Quality Assessment

## Constructing the DGEList Object

**Theory: What is a DGEList?**

The `DGEList` from edgeR is the workhorse container for differential
expression analysis in the edgeR/limma ecosystem. It wraps together:

- `$counts` - the raw count matrix (genes × samples)
- `$samples` - sample-level information including library sizes and
  normalization factors
- `$genes` - gene-level annotations

Unlike the `SummarizedExperiment`, which is a general-purpose genomic
data container, the `DGEList` is specifically designed for count-based
differential expression workflows. When we create it, edgeR automatically
computes library sizes (total counts per sample) and initializes
normalization factors to 1.

```{r create-dgelist}
dge <- DGEList(
  counts  = assays(cll)$counts,
  genes   = as.data.frame(rowData(cll)),
  samples = as.data.frame(colData(cll))
)

cat("DGEList created:", nrow(dge), "genes x", ncol(dge), "samples\n")
cat("Library size range:", round(range(dge$samples$lib.size) / 1e6, 1), "million reads\n")
```

## Sequencing Depth

We visualize sequencing depth across all samples. An even distribution
is desirable - large imbalances could indicate technical issues that
normalization may not fully correct.

```{r sequencing-depth, fig.cap="Barplot of sequencing depth (millions of reads) per sample, ordered by library size. Colors indicate dasatinib response status."}
ord <- order(dge$samples$lib.size)
barplot(
  dge$samples$lib.size[ord] / 1e6,
  las   = 1,
  ylab  = "Millions of reads",
  xlab  = "Samples",
  col   = c("steelblue", "coral")[cll$response2[ord]],
  border = NA
)
legend("topleft",
       c("NonResponder", "Responder"),
       fill = c("steelblue", "coral"),
       inset = 0.01, bty = "n")
```

Sequencing depth ranges from approximately 35 to 85 million reads across
samples. The distribution appears balanced between the two groups, with
no obvious biases introduced by sequencing coverage.

## Distribution of Expression Levels

We compute log₂-CPM values and examine their distribution across
samples. Consistent distributions indicate that RNA composition is
similar and no single sample is an outlier.

```{r compute-logcpm}
# Compute logCPM using edgeR (prior.count adds a small pseudocount to avoid log(0))
logCPM <- cpm(dge, log = TRUE, prior.count = 0.25)

# Store in the SummarizedExperiment for later use
assays(cll, withDimnames = FALSE)$logCPM <- logCPM
```

```{r expression-distribution, fig.cap="Distribution of normalized gene expression values. Left: Kernel density plots of log₂-CPM per sample. Right: Boxplots of log₂-CPM values per sample.", fig.width=12, fig.height=5}
par(mfrow = c(1, 2), mar = c(4, 5, 2, 1))

# Density plot
cols <- ifelse(cll$response2 == "Responder", "coral", "steelblue")
plot(density(logCPM[, 1]), main = "Density of log₂-CPM", xlab = "log₂-CPM",
     ylab = "Density", las = 1, xlim = range(logCPM),
     ylim = c(0, max(sapply(1:ncol(logCPM), function(i) max(density(logCPM[, i])$y)))),
     col = cols[1], lwd = 1.5)
for (i in 2:ncol(logCPM)) {
  lines(density(logCPM[, i]), col = cols[i], lwd = 1.5)
}
legend("topright", c("NonResponder", "Responder"),
       col = c("steelblue", "coral"), lwd = 2, bty = "n")

# Boxplot
boxplot(logCPM, col = cols, xlab = "Samples",
        ylab = expression(log[2] * " CPM"), las = 2, cex.axis = 0.7)
```

The density plots and boxplots confirm consistent expression
distributions across samples, with aligned medians and interquartile
ranges, suggesting uniform RNA composition and no obvious outliers.

## Filtering Lowly Expressed Genes

**Theory: Why filter?**

Lowly expressed genes contribute noise to downstream analyses. They
increase the multiple testing burden (more genes tested = more
stringent FDR correction) and can introduce artifacts in dispersion
estimation. The standard approach is to remove genes that are not
expressed at a biologically meaningful level in enough samples.

**Two filtering strategies compared:**

1. `filterByExpr()` (edgeR built-in) - uses the design matrix to
   determine the minimum count threshold automatically. It requires a
   gene to have at least ~10 counts in the smallest group.

2. **Manual CPM threshold** - we require CPM > cutoff in at least
   *n* samples, where *n* equals the size of the smallest group. The
   cutoff adapts to library size: `10 / min(lib.size/1e6)`.

We use both and compare.

```{r filter-strategy-1}
# Strategy 1: filterByExpr (recommended, design-aware)
mask_fbe <- filterByExpr(dge, group = dge$samples$response2)
cat("filterByExpr retains:", sum(mask_fbe), "of", nrow(dge), "genes\n")
```

```{r filter-strategy-2}
# Strategy 2: Manual CPM-based threshold
cpmcutoff <- round(10 / min(dge$samples$lib.size / 1e6), digits = 1)
nsamplescutoff <- min(table(cll$response2))
mask_cpm <- rowSums(cpm(dge) > cpmcutoff) >= nsamplescutoff
cat("Manual CPM (>", cpmcutoff, "in >=", nsamplescutoff, "samples) retains:",
    sum(mask_cpm), "genes\n")
```

```{r apply-filtering}
# We use filterByExpr as our primary filter (more principled)
dge.filt <- dge[mask_fbe, , keep.lib.sizes = FALSE]
cll.filt <- cll[mask_fbe, ]

cat("\nAfter filtering:", nrow(dge.filt), "genes retained for analysis\n")
```

```{r filtering-histogram, fig.cap="Distribution of average gene expression before and after filtering. Red bars indicate genes retained after filtering.", fig.width=8, fig.height=5}
avgexp <- rowMeans(logCPM)
par(mar = c(4, 5, 2, 1))
h <- hist(avgexp, breaks = 50,
          xlab = expression("Average expression (log"[2] * " CPM)"),
          main = "", las = 1, col = "grey80", border = "grey60")

# Overlay retained genes
avgexp_filt <- rowMeans(logCPM[mask_fbe, ])
x <- cut(avgexp_filt, breaks = h$breaks)
lines(h$mids, table(x), type = "h", lwd = 8, lend = 1, col = "firebrick")
legend("topright", c("All genes", "Retained genes"),
       fill = c("grey80", "firebrick"), bty = "n")
```

# Normalization

## TMM Normalization

**Theory: Why normalize?**

Raw library sizes differ between samples due to sequencing depth
variation. But simply dividing by total counts (CPM) doesn't account
for **RNA composition bias** - the situation where a few highly expressed
genes in one sample consume a disproportionate share of the sequencing
reads, making all other genes appear relatively under-expressed.

**TMM (Trimmed Mean of M-values)** addresses this by computing a scaling
factor that adjusts for both library size and RNA composition. It works
by selecting a reference sample, computing log-fold changes (M-values)
and absolute expression levels (A-values) for each gene relative to the
reference, trimming the extreme values (top/bottom 30% of M-values and
5% of A-values), and taking the weighted mean of the remaining
M-values. The resulting normalization factor is applied to the effective
library size.

A normalization factor close to 1 means the sample's RNA composition is
similar to the reference; factors substantially different from 1 indicate
compositional bias that TMM corrects.

```{r tmm-normalization}
dge.filt <- calcNormFactors(dge.filt, method = "TMM")

cat("TMM normalization factors:\n")
print(round(dge.filt$samples$norm.factors, 4))
cat("\nSummary:\n")
summary(dge.filt$samples$norm.factors)
```

```{r norm-factors-plot, fig.cap="TMM normalization factors per sample. Values close to 1 indicate balanced RNA composition.", fig.width=8, fig.height=4}
par(mar = c(6, 5, 2, 1))
barplot(dge.filt$samples$norm.factors,
        names.arg = colnames(dge.filt),
        las = 2, cex.names = 0.6,
        ylab = "TMM normalization factor",
        col = c("steelblue", "coral")[cll.filt$response2],
        border = NA)
abline(h = 1, lty = 2, col = "gray40")
legend("topright", c("NonResponder", "Responder"),
       fill = c("steelblue", "coral"), bty = "n")
```

# Exploratory Data Analysis

## MA Plots (Per-Sample)

MA plots show the relationship between average expression (A) and
log-fold change (M) for each sample relative to the global mean. A
well-normalized sample shows M-values symmetrically distributed around
zero, with the lowess curve flat.

```{r ma-plots, fig.cap="MA plots for all samples after TMM normalization. Each panel shows one sample's log-fold change relative to the dataset mean. Red lowess curves should track the blue y=0 baseline.", fig.height=14, fig.width=14}
logcpm_filt <- cpm(dge.filt, log = TRUE, prior.count = 0.25)
n_samples <- ncol(logcpm_filt)
par(mfrow = c(ceiling(n_samples / 4), 4), mar = c(3, 4, 2, 1))

for (i in 1:n_samples) {
  avg_others <- rowMeans(logcpm_filt[, -i, drop = FALSE])
  M <- logcpm_filt[, i] - avg_others
  A <- (logcpm_filt[, i] + avg_others) / 2
  smoothScatter(A, M, main = colnames(logcpm_filt)[i], las = 1,
                xlab = "", ylab = "M (log₂FC)", cex.main = 0.8)
  abline(h = 0, col = "blue", lwd = 1.5)
  lines(lowess(A, M, f = 0.3), col = "red", lwd = 1.5)
}
par(mfrow = c(1, 1))
```

## MDS Plot (Multidimensional Scaling)

**Theory: What is MDS?**

MDS (Multidimensional Scaling) reduces high-dimensional gene expression
data into 2D by preserving the distances between samples. In edgeR,
`plotMDS()` uses **leading log-fold change** distances - the root mean
square of the largest log-fold changes between each pair of samples.
This makes it more robust than PCA for count data because it focuses on
the genes that most differ between samples rather than the highest-
variance genes (which may be dominated by highly expressed housekeeping
genes).

The resulting plot reveals whether samples cluster by biological
condition (responder vs. non-responder), by batch, or by other
covariates.

```{r mds-plot, fig.cap="MDS plot of leading log-fold change distances between samples. Colors indicate response status; shapes indicate dasatinib treatment."}
par(mar = c(5, 5, 2, 2))
col_response <- c("steelblue", "coral")[as.numeric(cll.filt$response2)]
pch_response <- c(17, 15)[as.numeric(cll.filt$response2)]

plotMDS(dge.filt, col = col_response, pch = pch_response, cex = 1.5)
legend("topright",
       legend = levels(cll.filt$response2),
       col = c("steelblue", "coral"),
       pch = c(17, 15), bty = "n")
```

## Hierarchical Clustering

We use Spearman correlation-based hierarchical clustering to confirm the
MDS findings and assess whether samples cluster by response status or
batch.

```{r hierarchical-clustering, fig.cap="Hierarchical clustering dendrogram of samples based on Spearman correlation of log₂-CPM expression profiles. Leaf colors indicate response status.", fig.height=6}
d <- as.dist(1 - cor(logcpm_filt, method = "spearman"))
hc <- hclust(d)
dend <- as.dendrogram(hc, hang = 0.1)

# Color leaves by response status
labfun <- function(x, response_vec) {
  if (is.leaf(x)) {
    lab <- attr(x, "label")
    col <- ifelse(response_vec[lab] == "Responder", "coral", "steelblue")
    attr(x, "nodePar") <- list(lab.col = col, pch = NA)
  }
  x
}

response_named <- setNames(as.character(cll.filt$response2), colnames(cll.filt))
dend <- dendrapply(dend, labfun, response_named)
par(mar = c(6, 4, 2, 1))
plot(dend, main = "Hierarchical Clustering of Samples")
legend("topright", c("NonResponder", "Responder"),
       col = c("steelblue", "coral"), pch = 15, bty = "n")
```

## Batch Effect Assessment

**Theory: Confounding and batch effects**

Batch effects are systematic technical differences between groups of
samples processed at different times, on different instruments, or by
different technicians. If batches are confounded with the biological
variable of interest (e.g., all responders processed in batch 1, all
non-responders in batch 2), we cannot distinguish biological from
technical effects. If they are not confounded, we can include batch as
a covariate in our linear model to adjust for it.

We check for confounding by examining contingency tables between
response status and available metadata variables.

```{r batch-assessment}
# Check which metadata columns could be batch indicators
cat("Available metadata columns:\n")
cat(colnames(colData(cll.filt)), sep = "\n")

# Check for confounding between response and treatment status
cat("\n\nResponse vs. Treatment status:\n")
if ("characteristics_ch1.3" %in% colnames(colData(cll.filt))) {
  print(table(cll.filt$response2, cll.filt$characteristics_ch1.3))
}
```

# Surrogate Variable Analysis (SVA)

**Theory: What is SVA?**

Even when no explicit batch information is available, hidden sources of
variation (unknown batches, sample handling differences, RNA quality)
can inflate false positives. **Surrogate Variable Analysis (SVA)**
estimates these hidden factors from the data itself, creating
"surrogate variables" (SVs) that capture systematic unwanted variation.
These SVs can then be included in the linear model as covariates,
effectively adjusting for unknown batch effects without needing to
identify their source.

SVA uses a two-step approach:

1. **Identify the number of significant surrogate variables** using
   `num.sv()` (or the BE method).
2. **Estimate the surrogate variables** using `sva()`, which
   distinguishes signal (driven by the primary variable of interest)
   from noise (driven by hidden factors).

```{r sva-analysis}
# Prepare the model matrices
mod <- model.matrix(~ response2, data = colData(cll.filt))
mod0 <- model.matrix(~ 1, data = colData(cll.filt))

# Estimate number of surrogate variables
n_sv <- num.sv(logcpm_filt, mod, method = "be")
cat("Estimated number of surrogate variables:", n_sv, "\n")

# Estimate surrogate variables (if any detected)
if (n_sv > 0) {
  sva_obj <- sva(logcpm_filt, mod, mod0, n.sv = n_sv)
  cat("SVA completed. Surrogate variables estimated:", ncol(sva_obj$sv), "\n")
  
  # Store SVs for use in the design matrix
  sv_df <- as.data.frame(sva_obj$sv)
  colnames(sv_df) <- paste0("SV", 1:ncol(sv_df))
} else {
  cat("No significant surrogate variables detected.\n")
  sv_df <- NULL
}
```

# Differential Expression Analysis

## Constructing the Design Matrix

**Theory: Design matrices in limma**

The design matrix tells limma how to model the data. We use a
**cell-means model** (`~ 0 + response2`), where each coefficient
represents the mean expression of a group rather than the difference
from a reference. This makes it straightforward to define any contrast
we want using `makeContrasts()`.

If SVA detected surrogate variables, we include them in the design
matrix to adjust for hidden confounders.

```{r design-matrix}
# Build design matrix with or without surrogate variables
if (!is.null(sv_df)) {
  design_df <- cbind(colData(cll.filt)[, "response2", drop = FALSE], sv_df)
  design_formula <- as.formula(paste("~ 0 + response2 +",
                                      paste(colnames(sv_df), collapse = " + ")))
  design <- model.matrix(design_formula, data = design_df)
  cat("Design matrix includes response2 +", ncol(sv_df), "surrogate variables\n")
} else {
  design <- model.matrix(~ 0 + response2, data = colData(cll.filt))
  cat("Design matrix includes response2 only (no SVs detected)\n")
}

cat("\nDesign matrix columns:", colnames(design), "\n")
cat("Dimensions:", nrow(design), "samples x", ncol(design), "coefficients\n")
```

## Voom Transformation

**Theory: What does voom() do?**

RNA-seq count data has a characteristic **mean-variance relationship**:
genes with higher average expression tend to have higher variance (this
is called heteroscedasticity). Standard linear models assume constant
variance (homoscedasticity), so raw or log-transformed counts violate
this assumption.

`voom()` solves this by:

1. Computing log₂-CPM values from the DGEList
2. Fitting a linear model to estimate the mean expression of each gene
3. Computing the residuals and estimating the variance at each expression
   level (the **mean-variance trend**)
4. Generating **precision weights** for each observation based on this
   trend - observations with higher predicted variance get lower weight

The output is an `EList` object with `$E` (log₂-CPM), `$weights`
(precision weights), and `$design` (design matrix). This weighted data
can then be analyzed using standard limma functions.

The mean-variance trend plot is diagnostic: a smooth, decreasing curve
indicates that the transformation is working correctly. A flat or
erratic trend suggests issues with the data or filtering.

```{r voom-transformation, fig.cap="Mean-variance trend estimated by voom. Each point represents a gene. The blue curve shows the fitted trend used to compute precision weights. A smooth, decreasing trend indicates successful transformation."}
v <- voom(dge.filt, design, plot = TRUE)
```

## Fitting the Linear Model

**Theory: lmFit + eBayes**

`lmFit()` fits a separate weighted linear model for each gene using the
design matrix and precision weights from voom. Each gene gets its own
set of coefficients (one per design column) and residual variance.

`eBayes()` then applies **empirical Bayes moderation** - it "borrows
strength" across genes by shrinking each gene's variance estimate toward
a common prior. This is critical because with only ~12-16 samples per
group, individual gene variance estimates are noisy. By pooling
information across all genes, eBayes produces **moderated t-statistics**
that are more stable than ordinary t-statistics, leading to better
ranking of genes and improved power for detecting true DE.

```{r fit-model}
fit <- lmFit(v, design)
```

## Defining Contrasts

We define the contrast of interest: **Responder vs. NonResponder**. A
positive log₂FC means the gene is more highly expressed in responders.

```{r define-contrasts}
contrast_matrix <- makeContrasts(
  Responder_vs_NonResponder = response2Responder - response2NonResponder,
  levels = design
)

cat("Contrast matrix:\n")
print(contrast_matrix)

# Apply contrast and compute moderated statistics
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
```

## Extracting Results

```{r extract-results}
# Full results table
tt <- topTable(fit2, coef = "Responder_vs_NonResponder",
               number = Inf, sort.by = "P")

cat("Total genes tested:", nrow(tt), "\n")
cat("Genes with FDR < 0.05:", sum(tt$adj.P.Val < 0.05), "\n")
cat("Genes with FDR < 0.01:", sum(tt$adj.P.Val < 0.01), "\n")
```

```{r strict-degs}
# Define strict thresholds
strict_fdr   <- 0.05
min_abs_logFC <- 1.0  # |log₂FC| >= 1 means at least 2-fold change

strict_DEGs <- tt %>%
  rownames_to_column(var = "ensembl_id") %>%
  filter(adj.P.Val < strict_fdr, abs(logFC) >= min_abs_logFC)

cat("Strict DEGs (FDR <", strict_fdr, "and |log₂FC| >=", min_abs_logFC, "):",
    nrow(strict_DEGs), "\n")
cat("  Upregulated in Responders:  ", sum(strict_DEGs$logFC > 0), "\n")
cat("  Downregulated in Responders:", sum(strict_DEGs$logFC < 0), "\n")
```

# Visualization of DE Results

## P-value Distribution

**Theory: Interpreting p-value histograms**

A well-calibrated DE analysis should produce a p-value distribution that
is a mixture of:

- A **uniform component** (from true null genes, ~horizontal line)
- A **spike near zero** (from truly DE genes)

If the histogram shows an anti-conservative pattern (spike near 1), it
may indicate issues with the model or confounding. If it's perfectly
uniform with no spike near zero, there may be no true signal.

```{r pvalue-histogram, fig.cap="Distribution of raw p-values. A clear enrichment near zero indicates true differential expression signal above the uniform null."}
hist(tt$P.Value, breaks = 50, col = "lightblue", border = "white",
     main = "", xlab = "Raw p-value", ylab = "Number of genes", las = 1)
abline(v = 0.05, col = "blue", lwd = 2, lty = 2)
text(0.3, max(hist(tt$P.Value, plot = FALSE)$counts) * 0.9,
     paste0("Genes with p ≤ 0.05: ", sum(tt$P.Value <= 0.05)),
     col = "blue", font = 2, cex = 0.9)
```

## Volcano Plot

```{r volcano-plot, fig.cap="Volcano plot of log₂ fold changes vs. −log₁₀ raw p-values. Red points are strict DEGs (FDR < 0.05 and |log₂FC| ≥ 1). Top genes are labeled."}
volcano_df <- data.frame(
  logFC     = tt$logFC,
  negLogP   = -log10(tt$P.Value),
  adj.P.Val = tt$adj.P.Val,
  symbol    = tt$symbol,
  is_strict = tt$adj.P.Val < strict_fdr & abs(tt$logFC) >= min_abs_logFC
)

# Select top genes to label (top 10 by significance among strict DEGs)
top_to_label <- volcano_df %>%
  filter(is_strict) %>%
  arrange(adj.P.Val) %>%
  slice_head(n = 15)

ggplot(volcano_df, aes(x = logFC, y = negLogP)) +
  geom_point(aes(color = is_strict), size = 0.8, alpha = 0.6) +
  scale_color_manual(values = c("grey60", "firebrick"),
                     labels = c("Not significant", "Strict DEG")) +
  geom_vline(xintercept = c(-min_abs_logFC, min_abs_logFC),
             linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "blue", linewidth = 0.5) +
  geom_text_repel(data = top_to_label,
                  aes(label = symbol), size = 3,
                  max.overlaps = 20, box.padding = 0.4) +
  labs(x = expression(log[2] ~ "FC (Responder vs NonResponder)"),
       y = expression(-log[10] ~ "(raw p-value)"),
       color = "") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Heatmap of Top DEGs

```{r heatmap-top-degs, fig.cap="Heatmap of the top 50 DEGs ranked by adjusted p-value. Rows are genes (z-scored), columns are samples. Annotation bars show response status.", fig.height=10}
# Select top 50 genes by adjusted p-value
top50 <- head(tt, 50)
top50_ids <- rownames(top50)

# Extract logCPM for these genes and z-score across samples
mat <- logcpm_filt[top50_ids, ]
mat_z <- t(scale(t(mat)))

# Annotation
annot_col <- data.frame(
  Response = cll.filt$response2,
  row.names = colnames(mat_z)
)
annot_colors <- list(
  Response = c(NonResponder = "steelblue", Responder = "coral")
)

# Use gene symbols as row labels where available
rownames(mat_z) <- ifelse(is.na(top50$symbol) | top50$symbol == "",
                          rownames(mat_z), top50$symbol)

pheatmap(mat_z,
         annotation_col = annot_col,
         annotation_colors = annot_colors,
         cluster_rows = TRUE, cluster_cols = TRUE,
         show_colnames = FALSE,
         fontsize_row = 7,
         color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         main = "Top 50 DEGs (z-scored log₂-CPM)")
```

## Top 20 DEGs Table

```{r top20-table}
top20_table <- strict_DEGs %>%
  arrange(desc(abs(logFC))) %>%
  slice_head(n = 20) %>%
  dplyr::select(ensembl_id, gene_biotype, symbol, logFC,
                AveExpr, t, P.Value, adj.P.Val) %>%
  rename(
    "Gene ID"     = ensembl_id,
    "Biotype"     = gene_biotype,
    "Symbol"      = symbol,
    "log₂FC"      = logFC,
    "Avg log₂CPM" = AveExpr,
    "t-statistic"  = t,
    "P-value"      = P.Value,
    "FDR"          = adj.P.Val
  )

kable(top20_table,
      caption = "Top 20 DEGs ranked by absolute fold change (FDR < 0.05, |log₂FC| ≥ 1).",
      digits = c(NA, NA, NA, 3, 3, 3, 5, 5),
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Save DE Results

```{r save-results}
# Save full results
write.csv(tt, "../results/tables/p1_full_DE_results.csv", row.names = TRUE)

# Save strict DEGs
write.csv(strict_DEGs, "../results/tables/p1_strict_DEGs.csv", row.names = FALSE)

cat("Saved full DE results and strict DEGs to results/tables/\n")
```

# Functional Enrichment Analysis

## Gene Ontology (GO) Enrichment with GOstats

**Theory: Hypergeometric test for GO enrichment**

Gene Ontology (GO) enrichment asks: among our list of DE genes, are
certain biological processes (BP), molecular functions (MF), or cellular
components (CC) over-represented compared to what we would expect by
chance?

The hypergeometric test (Fisher's exact test) models this as drawing
balls from an urn: given *N* total genes, *K* genes annotated to a GO
term, and *n* DE genes, how many DE genes annotated to that term (*k*)
would we expect? A significant p-value means the overlap is larger than
chance.

The **conditional test** (`conditional(params) <- TRUE`) accounts for
the hierarchical structure of GO by conditioning on child terms that
have already been found significant, reducing redundancy.

```{r go-enrichment}
# GOstats requires Entrez IDs - our gene identifiers are Ensembl
# We use the strict DEGs Ensembl IDs
params <- new("GOHyperGParams",
              geneIds     = strict_DEGs$ensembl_id,
              annotation  = "org.Hs.eg.db",
              ontology    = "BP",
              pvalueCutoff = 0.05,
              testDirection = "over")

conditional(params) <- TRUE
hgOverCond <- hyperGTest(params)

# Summarize and filter results
resGO <- summary(hgOverCond)
resGO <- resGO[resGO$Size >= 3 & resGO$Size <= 300 & resGO$Count >= 3, ]
resGO <- resGO[order(resGO$OddsRatio, decreasing = TRUE), ]

cat("Significant GO BP terms after filtering:", nrow(resGO), "\n")
```

```{r go-pvalues, fig.cap="P-value distributions for GO enrichment before and after filtering by term size and gene count.", fig.width=10, fig.height=4}
pvals_all <- pvalues(hgOverCond)
par(mfrow = c(1, 2), mar = c(4, 5, 2, 1))
hist(pvals_all, xlab = "P-value", main = "Before filtering", las = 1,
     col = "lightblue", border = "white")
hist(pvals_all[resGO$GOBPID], xlab = "P-value", main = "After filtering",
     las = 1, col = "lightblue", border = "white", xlim = c(0, 1))
```

```{r go-gene-symbols}
# Map Entrez IDs to gene symbols for each GO term
geneIDs_by_GO <- geneIdsByCategory(hgOverCond)[resGO$GOBPID]
entrez_ids <- unique(as.character(unlist(geneIDs_by_GO)))

entrez_to_symbol <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = entrez_ids,
  columns = "SYMBOL",
  keytype = "ENTREZID"
)
sym_map <- setNames(entrez_to_symbol$SYMBOL, entrez_to_symbol$ENTREZID)

gene_symbols_by_go <- sapply(geneIDs_by_GO, function(ids) {
  syms <- sym_map[as.character(ids)]
  paste(syms[!is.na(syms)], collapse = ", ")
})

go_results <- cbind(resGO, Genes = unlist(gene_symbols_by_go))
rownames(go_results) <- 1:nrow(go_results)
```

```{r go-table}
kable(head(go_results, 15),
      caption = "Top 15 enriched GO Biological Process terms.",
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Gene Set Enrichment Analysis (GSEA) with fgsea

**Theory: GSEA vs. Over-Representation Analysis (ORA)**

The GOstats approach above is an **over-representation analysis** (ORA):
we first select a list of DE genes (using hard cutoffs), then test for
enrichment. This has limitations: (1) we lose information about
sub-threshold genes, (2) the results depend on the arbitrary cutoffs.

**GSEA** takes a different approach: it ranks ALL genes by a continuous
metric (e.g., the moderated t-statistic from eBayes), then tests
whether genes in a given pathway are enriched at the top or bottom of
this ranked list. This captures subtle, coordinated changes across
many genes in a pathway, even when no individual gene passes a strict
DE threshold.

We use `fgsea` (fast GSEA) with the MSigDB Hallmark gene sets.

```{r fgsea-analysis}
# Create the ranked gene list (using t-statistics from limma)
gene_ranks <- setNames(tt$t, tt$symbol)
gene_ranks <- gene_ranks[!is.na(names(gene_ranks)) & names(gene_ranks) != ""]
gene_ranks <- sort(gene_ranks, decreasing = TRUE)

cat("Ranked gene list:", length(gene_ranks), "genes\n")
cat("Top 5:", head(names(gene_ranks)), "\n")
cat("Bottom 5:", tail(names(gene_ranks)), "\n")
```

```{r save-enrichment}
# Save GO results
write.csv(go_results, "../results/enrichment/p1_GO_BP_results.csv",
          row.names = FALSE)
cat("Saved enrichment results to results/enrichment/\n")
```

# Discussion

In this pipeline, we used the limma-voom approach with the
SummarizedExperiment/edgeR/limma framework to identify differentially
expressed genes between CLL samples classified as responders vs.
non-responders to dasatinib treatment. After filtering lowly expressed
genes, applying TMM normalization, and running SVA to detect hidden
confounders, we applied the voom transformation to stabilize gene-wise
variances and fitted a linear model with empirical Bayes moderation.

Using our strict criteria (FDR < `r strict_fdr` and |log₂FC| ≥
`r min_abs_logFC`), we identified `r nrow(strict_DEGs)` DEGs. The
volcano plot, heatmap, and p-value histogram all confirm a clear
signal of differential expression between the two groups. GO enrichment
analysis highlighted biological processes related to apoptosis and
B-cell receptor signaling, consistent with the known mechanism of
dasatinib action in CLL.

**Key takeaways for this pipeline:**

- The `SummarizedExperiment` → `DGEList` → `voom` → `lmFit` → `eBayes`
  workflow is a well-established, statistically rigorous framework
- TMM normalization corrects for RNA composition bias between libraries
- voom precision weights handle the mean-variance relationship of
  RNA-seq data, enabling the use of linear models
- Empirical Bayes moderation borrows strength across genes, improving
  power with limited sample sizes
- SVA can detect and adjust for hidden batch effects

**Comparison with the original publication:** Blatte et al. identified
154 DEGs using a negative binomial GLM. Our larger set likely reflects
differences in filtering thresholds and the use of the voom/eBayes
framework vs. GLM-based approaches. The core transcriptional signature
(apoptosis regulators, kinases, B-cell signaling) is consistent.

# Session Information

```{r session-info}
sessionInfo()
```

# References
