---
title: "Pipeline 2: Bulk RNA-seq Analysis of CLL Dasatinib Response"
subtitle: "Seurat Object + DESeq2 Approach"
author:
- name: "Samuel Escudero, Ivon Sanchez, Karim Hamed"
  affiliation: Universitat Pompeu Fabra
  email: samuel.escudero01@estudiant.upf.edu
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_captions: yes
    code_folding: show
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "",
  fig.align   = "center",
  fig.width   = 8,
  fig.height  = 6,
  fig.wide    = TRUE,
  cache       = FALSE,
  message     = FALSE,
  warning     = FALSE
)
```

# Preliminary: From FASTQ to Count Matrix

Before we begin our differential expression analysis, it is important to
understand how the raw sequencing data (FASTQ files) are processed to
produce the count matrix that we use as input. This section describes the
upstream steps that precede our R-based analysis. In our case, the
GSE151159 dataset was already processed and the count matrix is
available as an RDS file. However, if we were starting from raw FASTQ
files, we would need to perform the following steps.

## The upstream RNA-seq workflow

The general workflow from raw reads to a count matrix involves three
major steps:

1. **Quality control of raw reads** (FastQC, MultiQC)
2. **Alignment to a reference genome** (STAR) or pseudo-alignment
   to a transcriptome (Salmon, kallisto)
3. **Quantification** (featureCounts for genome-aligned BAMs, or
   built-in quantification in Salmon/kallisto)

## Option A: Genome alignment with STAR + featureCounts

STAR (Spliced Transcripts Alignment to a Reference) is a fast and
accurate splice-aware aligner designed specifically for RNA-seq. It
maps reads to the genome while correctly handling intron-spanning
reads (spliced alignments). After alignment, featureCounts (from the
Subread package) counts how many reads overlap each annotated gene.

```bash
# ---- STEP 0: Quality control of raw reads ----
# FastQC generates per-sample HTML reports with metrics like
# per-base quality scores, adapter content, and GC distribution
fastqc -o qc_results/ -t 8 raw_data/*.fastq.gz

# MultiQC aggregates all FastQC reports into a single summary
multiqc qc_results/ -o qc_results/multiqc/

# ---- STEP 1: Build the STAR genome index (run once) ----
# We need the reference genome FASTA and the GTF annotation file.
# For human data, we download these from ENSEMBL or GENCODE.
# The sjdbOverhang parameter should be (read length - 1).
# For 150bp reads, use 149; for 100bp reads, use 99.

STAR \
  -runMode genomeGenerate \
  -genomeDir star_index/ \
  -genomeFastaFiles Homo_sapiens.GRCh38.dna.primary_assembly.fa \
  -sjdbGTFfile Homo_sapiens.GRCh38.109.gtf \
  -sjdbOverhang 149 \
  -runThreadN 8

# ---- STEP 2: Align reads to the genome ----
# For each sample, we run STAR in alignment mode.
# Input: paired-end FASTQ files (R1 and R2)
# Output: a sorted BAM file with aligned reads

for sample in sample1 sample2 sample3; do
  STAR \
    -genomeDir star_index/ \
    -readFilesIn raw_data/${sample}_R1.fastq.gz \
                 raw_data/${sample}_R2.fastq.gz \
    -readFilesCommand zcat \
    -outSAMtype BAM SortedByCoordinate \
    -outFileNamePrefix aligned/${sample}_ \
    -runThreadN 8 \
    -quantMode GeneCounts
done

# ---- STEP 3: Count reads per gene with featureCounts ----
# featureCounts assigns aligned reads to genomic features (genes)
# defined in the GTF annotation file.
# -p: paired-end mode
# -s 2: reverse-stranded library (most common for Illumina)
# -T 8: use 8 threads
# -a: GTF annotation file
# -o: output count matrix

featureCounts \
  -p \
  -s 2 \
  -T 8 \
  -a Homo_sapiens.GRCh38.109.gtf \
  -o counts/gene_counts.txt \
  aligned/*_Aligned.sortedByCoord.out.bam
```

The output `gene_counts.txt` is a tab-separated file with gene IDs in
rows and samples in columns. We would then import this into R using
`read.table()` and construct our analysis objects.

## Option B: Pseudo-alignment with Salmon

Salmon takes a fundamentally different approach: instead of aligning
reads to the genome, it performs "pseudo-alignment" directly to the
transcriptome. This is much faster (minutes instead of hours) and does
not require a genome index, only a transcriptome index. Salmon also
applies bias correction models (GC bias, positional bias, fragment
length) that improve quantification accuracy.

```bash
# ---- STEP 1: Build the Salmon transcriptome index (run once) ----
# We use the cDNA FASTA from ENSEMBL (transcriptome, not genome)

salmon index \
  -t Homo_sapiens.GRCh38.cdna.all.fa.gz \
  -i salmon_index \
  -p 8

# ---- STEP 2: Quantify each sample ----
# Salmon performs pseudo-alignment and quantification in one step
# -l A: automatically detect library type
# -1 and -2: paired-end FASTQ files
# -p: number of threads

for sample in sample1 sample2 sample3; do
  salmon quant \
    -i salmon_index \
    -l A \
    -1 raw_data/${sample}_R1.fastq.gz \
    -2 raw_data/${sample}_R2.fastq.gz \
    -p 8 \
    -o salmon_quant/${sample}
done

# ---- STEP 3: Import into R with tximeta ----
# Salmon produces per-sample quant.sf files with transcript-level
# counts and TPM values. We use tximeta in R to import these and
# summarize to gene level.
```

To import Salmon output into R, we would use the `tximeta` package:

```r
# In R:
library(tximeta)
library(SummarizedExperiment)

# Create a sample table linking sample names to quant.sf paths
coldata <- data.frame(
  names  = c("sample1", "sample2", "sample3"),
  files  = file.path("salmon_quant",
                     c("sample1", "sample2", "sample3"),
                     "quant.sf"),
  condition = c("NonResponder", "NonResponder", "Responder")
)

# Import and summarize to gene level
se <- tximeta(coldata)
gse <- summarizeToGene(se)

# 'gse' is now a SummarizedExperiment with gene-level counts
# ready for DESeq2 or edgeR analysis
```

## STAR vs. Salmon: Key differences

| Feature | STAR + featureCounts | Salmon |
|---------|---------------------|--------|
| Approach | Genome alignment | Transcriptome pseudo-alignment |
| Speed | Slower (hours) | Very fast (minutes) |
| Output | BAM + count matrix | Transcript-level counts + TPM |
| Memory | High (30+ GB for human) | Low (< 8 GB) |
| Splice-awareness | Yes (detects novel junctions) | No (uses known transcripts) |
| Bias correction | No built-in | Yes (GC, positional, fragment) |
| R import | read.table() or Rsubread | tximeta (preserves metadata) |
| Best for | Novel transcript discovery | Standard DE analysis |

For our CLL dataset (GSE151159), the count matrix was already generated
by the original authors and deposited at GEO. We load it directly as
an RDS file containing a SummarizedExperiment object.

# Introduction

## Biological context

Chronic lymphocytic leukemia (CLL) is a clinically heterogeneous
hematological malignancy characterized by diverse responses to targeted
treatments. Dasatinib, a second-generation tyrosine kinase inhibitor
(TKI), represents a potential therapeutic strategy for patients with CLL.
However, its clinical efficacy varies significantly between individuals.
Blatte et al. (2021) employed RNA sequencing (RNA-seq) to characterize
the transcriptional profiles of primary leukemia cells from
treatment-naive CLL patients classified into two groups based on in vitro
cytotoxicity assays: **responders** (dasatinib-sensitive) and
**nonresponders** (dasatinib-resistant). Their primary goal was to
discover gene expression signatures predictive of dasatinib sensitivity.

The RNAseq data generated by Blatte and colleagues were deposited at the
Gene Expression Omnibus (GEO), publicly accessible under accession
[GSE151159](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151159).

## Analytical approach: Seurat + DESeq2

This is **Pipeline 2** of our multi-approach analysis. We use a
**Seurat object** as our data container and **DESeq2** as our
differential expression framework. This pipeline is fully
self-contained and loads all data directly from the RDS file.

**Theory: The Seurat object**

Seurat was originally developed for single-cell RNA-seq (scRNA-seq)
analysis by the Satija Lab (Butler et al., 2018; Stuart et al., 2019).
However, its data structure is perfectly capable of handling bulk
RNA-seq data. A Seurat object stores:

- **Assays**: named slots containing expression matrices. The default
  `RNA` assay holds raw counts in `counts`, normalized data in `data`,
  and scaled data in `scale.data`.
- **Meta.data**: a data frame of sample-level (or cell-level in
  scRNA-seq) metadata, equivalent to `colData()` in
  SummarizedExperiment.
- **Reductions**: dimensionality reduction results (PCA, UMAP, tSNE).
- **Graphs**: nearest-neighbor graphs for clustering.

**Theory: DESeq2 vs. limma-voom**

While Pipeline 1 used `voom()` to transform counts into weighted
log-CPM values and then applied linear models, DESeq2 takes a
fundamentally different approach:

1. **Negative binomial model**: DESeq2 models the raw counts directly
   using a negative binomial (NB) distribution, which naturally handles
   the overdispersion (variance > mean) typical of RNA-seq data.

2. **Shrinkage of dispersion estimates**: DESeq2 estimates gene-wise
   dispersions and then shrinks them toward a fitted trend using an
   empirical Bayes procedure. This is conceptually similar to what
   limma's `eBayes()` does for variances, but operates on the
   dispersion parameter of the NB distribution.

3. **Wald test**: For each gene, DESeq2 tests whether the log2 fold
   change coefficient is significantly different from zero using a
   Wald test (ratio of the estimated LFC to its standard error).

4. **Log fold change shrinkage**: After testing, DESeq2 can apply
   LFC shrinkage using methods like `apeglm`, which produces more
   reliable effect size estimates for genes with low counts or high
   dispersion.

| Feature | limma-voom (Pipeline 1) | DESeq2 (Pipeline 2) |
|---------|------------------------|---------------------|
| Input | Transformed log-CPM + weights | Raw counts directly |
| Distribution | Normal (after voom) | Negative binomial |
| Variance stabilization | voom precision weights | NB dispersion shrinkage |
| Test | Moderated t-test | Wald test |
| LFC shrinkage | Not built-in | apeglm/ashr/normal |
| Speed | Faster | Slower (iterative fitting) |
| Best for | Large samples, complex designs | Small-moderate samples |

## Libraries

```{r load-libraries}
# Core packages
library(Seurat)               # Data container and visualization
library(DESeq2)               # Differential expression (NB model)
library(SummarizedExperiment)  # To read the RDS

# Batch effect detection
library(sva)                  # Surrogate Variable Analysis

# Visualization
library(ggplot2)              # Grammar of graphics
library(ggrepel)              # Non-overlapping labels
library(pheatmap)             # Clustered heatmaps
library(RColorBrewer)         # Color palettes

# Functional enrichment
library(GOstats)              # Hypergeometric GO testing
library(org.Hs.eg.db)         # Human gene annotations
library(AnnotationDbi)        # Annotation interface

# Data manipulation and reporting
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
```

# Data Import and Seurat Object Construction

## Loading the SummarizedExperiment

We load the `GSE151159.rds` file directly, extract the components we
need, and classify samples into responders and non-responders. This
pipeline is self-contained and does not depend on any exports from
Pipeline 1.

```{r load-data}
# Load the SummarizedExperiment directly from the RDS
cll <- readRDS("../data/GSE151159.rds")

cat("Loaded:", class(cll), "\n")
cat("Dimensions:", nrow(cll), "genes x", ncol(cll), "samples\n")

# Inspect the structure
cat("\nAssays available:", names(assays(cll)), "\n")
cat("Sample metadata columns:", colnames(colData(cll)), "\n")
cat("Gene annotation columns:", colnames(rowData(cll)), "\n")
```

## Sample Classification

We create a clean binary factor `response2` from the raw
`description.2` metadata field, then remove ambiguous samples labeled
"all".

```{r classify-samples}
# Examine raw descriptions
cat("Raw sample descriptions:\n")
table(colData(cll)$description.2)

# Create clean response classification
desc2 <- colData(cll)$description.2
response_raw <- ifelse(
  grepl("^responders_final", desc2), "Responder",
  ifelse(grepl("^nonresponders_final", desc2), "NonResponder", NA)
)

colData(cll)$response2 <- factor(
  response_raw,
  levels = c("NonResponder", "Responder")
)

cat("\nClean response classification:\n")
table(colData(cll)$response2, useNA = "ifany")

# Remove samples with undefined response (labeled "all")
keep_samples <- !is.na(colData(cll)$response2)
cll <- cll[, keep_samples]

cat("\nAfter removing ambiguous samples:", ncol(cll), "samples retained\n")
cat("  NonResponders:", sum(cll$response2 == "NonResponder"), "\n")
cat("  Responders:   ", sum(cll$response2 == "Responder"), "\n")
```

## Extracting Data Components

```{r extract-components}
# Extract count matrix, sample metadata, and gene metadata
counts_mat <- as.data.frame(assays(cll)$counts)
sample_meta <- as.data.frame(colData(cll))
gene_meta   <- as.data.frame(rowData(cll))

cat("Count matrix:", nrow(counts_mat), "genes x", ncol(counts_mat), "samples\n")
cat("\nFirst 5 gene annotations:\n")
head(gene_meta, 5)
```

## Creating the Seurat Object

**Theory: CreateSeuratObject internals**

When we call `CreateSeuratObject()`, Seurat performs several things
under the hood:

1. Creates an `Assay` object containing the count matrix in the
   `counts` layer
2. Computes `nCount_RNA` (total counts per sample, equivalent to
   library size) and `nFeature_RNA` (number of detected genes per
   sample)
3. Stores sample metadata in the `meta.data` slot
4. Sets the default assay to `RNA`

The `min.cells` and `min.features` parameters are filtering thresholds.
For bulk RNA-seq, `min.cells` controls how many samples must express a
gene for it to be retained. We set `min.cells = 3` as a light
pre-filter (genes must be detected in at least 3 samples).

```{r create-seurat}
# Ensure response2 is a factor
sample_meta$response2 <- factor(sample_meta$response2,
                                 levels = c("NonResponder", "Responder"))

# Create Seurat object
seurat_obj <- CreateSeuratObject(
  counts    = as.matrix(counts_mat),
  meta.data = sample_meta,
  project   = "CLL_Dasatinib",
  min.cells = 3,       # Gene must be in >= 3 samples
  min.features = 200   # Sample must have >= 200 detected genes
)

cat("Seurat object created:\n")
seurat_obj
```

```{r inspect-seurat}
# Inspect the structure
cat("Default assay:", DefaultAssay(seurat_obj), "\n")
cat("Assay layers:", Layers(seurat_obj), "\n")
cat("Metadata columns:", colnames(seurat_obj@meta.data), "\n\n")

# Check nCount and nFeature distributions
cat("Library size (nCount_RNA) range:",
    range(seurat_obj$nCount_RNA), "\n")
cat("Detected genes (nFeature_RNA) range:",
    range(seurat_obj$nFeature_RNA), "\n")
cat("Number of samples:", ncol(seurat_obj), "\n")
```

# Quality Control

## QC Metrics Visualization

In scRNA-seq, we would filter cells based on mitochondrial percentage,
doublet scores, and gene/UMI counts. For bulk RNA-seq, these metrics
are less critical because each "cell" is actually a bulk sample with
millions of reads. However, visualizing library size and gene detection
rates helps us identify potential outlier samples.

```{r qc-violin, fig.cap="Distribution of QC metrics across samples. Left: total read counts (library size). Right: number of detected genes per sample."}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA"),
        group.by = "response2",
        pt.size = 0.5,
        cols = c("steelblue", "coral")) +
  theme(axis.text.x = element_text(angle = 0))
```

```{r qc-scatter, fig.cap="Relationship between library size and number of detected genes, colored by response group. A strong linear relationship is expected for good-quality samples."}
FeatureScatter(seurat_obj,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA",
               group.by = "response2",
               cols = c("steelblue", "coral"))
```

## Sequencing Depth Barplot

```{r seq-depth, fig.cap="Barplot of sequencing depth (millions of reads) per sample, ordered by library size. Colors indicate dasatinib response status."}
ord <- order(seurat_obj$nCount_RNA)
barplot(
  seurat_obj$nCount_RNA[ord] / 1e6,
  las   = 1,
  ylab  = "Millions of reads",
  xlab  = "Samples",
  col   = c("steelblue", "coral")[seurat_obj$response2[ord]],
  border = NA
)
legend("topleft",
       c("NonResponder", "Responder"),
       fill = c("steelblue", "coral"),
       inset = 0.01, bty = "n")
```

# Normalization

## LogNormalize (Seurat default)

**Theory: Seurat's LogNormalize**

Seurat's default normalization uses a simple approach:

1. Divide each gene's count by the total counts for that sample
   (equivalent to calculating a proportion)
2. Multiply by a scale factor (default: 10,000)
3. Take the natural log (log1p, i.e., log(x + 1))

This is conceptually similar to computing log-CPM (counts per million),
but with a scale factor of 10,000 instead of 1,000,000. The formula is:

`normalized_value = log(count / total_counts * scale_factor + 1)`

This is simpler than TMM (which adjusts for RNA composition) but works
well in practice for many datasets. The normalized values are stored in
the `data` layer of the RNA assay.

```{r normalize}
seurat_obj <- NormalizeData(
  seurat_obj,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)

cat("Normalization complete.\n")
cat("Layers after normalization:", Layers(seurat_obj), "\n")
```

## Identify Highly Variable Features

**Theory: Why select variable features?**

For dimensionality reduction (PCA, UMAP), we want to focus on genes
that show meaningful variation across samples, not genes with constant
expression. Highly variable genes (HVGs) are those with the most
variance relative to their mean expression. Seurat uses the `vst`
(variance-stabilizing transformation) method to identify HVGs,
which models the relationship between mean and variance and selects
genes that deviate most from this trend.

In scRNA-seq, we typically select 2000-3000 HVGs. For bulk RNA-seq with
fewer samples, we can be more generous (3000-5000) to capture more
of the biological variation.

```{r find-variable-features, fig.cap="Highly variable genes identified by the VST method. Red points are the selected variable features. Top 10 are labeled."}
seurat_obj <- FindVariableFeatures(
  seurat_obj,
  selection.method = "vst",
  nfeatures = 3000
)

# Plot variable features
top10_hvg <- head(VariableFeatures(seurat_obj), 10)
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10_hvg, repel = TRUE)
plot2
```

## Scale Data

**Theory: ScaleData**

Scaling performs two operations per gene:

1. **Centers** the expression values (subtracts the mean across samples)
2. **Scales** by dividing by the standard deviation

After scaling, each gene has mean = 0 and SD = 1. This is essential for
PCA because without scaling, PCA would be dominated by highly expressed
genes simply because they have larger absolute values, not because they
carry more biological information. The scaled data is stored in the
`scale.data` layer.

```{r scale-data}
seurat_obj <- ScaleData(seurat_obj, features = rownames(seurat_obj))
cat("Data scaled. Layers:", Layers(seurat_obj), "\n")
```

# Dimensionality Reduction

## PCA

**Theory: PCA on bulk RNA-seq via Seurat**

PCA (Principal Component Analysis) finds the axes of maximum variance
in the data. Seurat's `RunPCA()` operates on the scaled data and
computes the top principal components. For bulk RNA-seq, PC1 and PC2
typically capture the dominant biological signal (e.g., treatment
response), while later PCs may capture batch effects or noise.

The **elbow plot** helps us decide how many PCs contain real signal:
we look for the "elbow" where the standard deviation drops sharply,
indicating that subsequent PCs capture mostly noise.

Note that in Pipeline 1 we used MDS (multidimensional scaling) with
leading log-fold change distances. PCA and MDS are conceptually similar
but not identical: MDS preserves sample distances directly, while PCA
finds the directions of maximum variance. For well-normalized bulk
RNA-seq, both should give similar sample groupings.

```{r run-pca}
seurat_obj <- RunPCA(seurat_obj,
                      features = VariableFeatures(seurat_obj),
                      npcs = 20)
```

```{r pca-plot, fig.cap="PCA of samples colored by dasatinib response status. Each point is one sample."}
DimPlot(seurat_obj, reduction = "pca", group.by = "response2",
        cols = c("steelblue", "coral"), pt.size = 3) +
  ggtitle("PCA: Responders vs NonResponders")
```

```{r elbow-plot, fig.cap="Elbow plot showing the standard deviation explained by each principal component. The elbow point suggests how many PCs carry meaningful signal."}
ElbowPlot(seurat_obj, ndims = 20)
```

```{r pca-loadings, fig.cap="Top genes contributing to PC1 and PC2. These genes drive the primary separation between samples."}
VizDimLoadings(seurat_obj, dims = 1:2, reduction = "pca")
```

## UMAP

**Theory: UMAP vs. PCA**

UMAP (Uniform Manifold Approximation and Projection) is a nonlinear
dimensionality reduction technique that preserves both local and global
structure better than PCA. While PCA is limited to linear relationships,
UMAP can capture complex, nonlinear patterns in the data. However, UMAP
is stochastic (results vary between runs) and distances in UMAP space
are not directly interpretable. For bulk RNA-seq with few samples
(~28), UMAP adds limited value over PCA, but it is standard practice in
Seurat workflows and provides a nice visualization.

**Important note on n.neighbors:** UMAP's default `n.neighbors = 30`
requires at least 31 samples. With our 28 samples, we must set
`n.neighbors` to a value smaller than the number of samples. We use
`n.neighbors = min(15, ncol(seurat_obj) - 1)` to adapt automatically.

```{r run-umap, fig.cap="UMAP projection of samples colored by response status."}
# Adapt n.neighbors to our small sample size
# Default is 30, but we only have 28 samples, so we reduce it
n_samples <- ncol(seurat_obj)
n_neighbors_safe <- min(15, n_samples - 1)

cat("Number of samples:", n_samples, "\n")
cat("Using n.neighbors:", n_neighbors_safe, "\n")

seurat_obj <- RunUMAP(seurat_obj,
                       dims = 1:10,
                       n.neighbors = n_neighbors_safe)

DimPlot(seurat_obj, reduction = "umap", group.by = "response2",
        cols = c("steelblue", "coral"), pt.size = 3) +
  ggtitle("UMAP: Responders vs NonResponders")
```

## Hierarchical Clustering

We use Spearman correlation-based hierarchical clustering as an
independent validation of the PCA and UMAP sample groupings.

```{r hierarchical-clustering, fig.cap="Hierarchical clustering dendrogram of samples based on Spearman correlation. Leaf colors indicate response status.", fig.height=6}
logcpm_seurat <- as.matrix(GetAssayData(seurat_obj, layer = "data"))
d <- as.dist(1 - cor(logcpm_seurat, method = "spearman"))
hc <- hclust(d)
dend <- as.dendrogram(hc, hang = 0.1)

# Color leaves by response status
labfun <- function(x, response_vec) {
  if (is.leaf(x)) {
    lab <- attr(x, "label")
    col <- ifelse(response_vec[lab] == "Responder", "coral", "steelblue")
    attr(x, "nodePar") <- list(lab.col = col, pch = NA)
  }
  x
}

response_named <- setNames(
  as.character(seurat_obj$response2),
  colnames(seurat_obj)
)
dend <- dendrapply(dend, labfun, response_named)
par(mar = c(6, 4, 2, 1))
plot(dend, main = "Hierarchical Clustering of Samples")
legend("topright", c("NonResponder", "Responder"),
       col = c("steelblue", "coral"), pch = 15, bty = "n")
```

# Surrogate Variable Analysis (SVA)

**Theory: What is SVA?**

Even when no explicit batch information is available, hidden sources of
variation (unknown batches, sample handling differences, RNA quality)
can inflate false positives. Surrogate Variable Analysis (SVA) estimates
these hidden factors from the data itself, creating "surrogate
variables" (SVs) that capture systematic unwanted variation. These SVs
can then be included as covariates in the DESeq2 model.

```{r sva-analysis}
# Compute logCPM for SVA input
logcpm_for_sva <- as.matrix(GetAssayData(seurat_obj, layer = "data"))

# Prepare model matrices
mod  <- model.matrix(~ response2, data = seurat_obj@meta.data)
mod0 <- model.matrix(~ 1, data = seurat_obj@meta.data)

# Estimate number of surrogate variables
n_sv <- num.sv(logcpm_for_sva, mod, method = "be")
cat("Estimated number of surrogate variables:", n_sv, "\n")

# Estimate surrogate variables (if any detected)
if (n_sv > 0) {
  sva_obj <- sva(logcpm_for_sva, mod, mod0, n.sv = n_sv)
  sv_df <- as.data.frame(sva_obj$sv)
  colnames(sv_df) <- paste0("SV", 1:ncol(sv_df))
  rownames(sv_df) <- colnames(seurat_obj)
  cat("SVA completed. Surrogate variables estimated:", ncol(sv_df), "\n")
} else {
  cat("No significant surrogate variables detected.\n")
  sv_df <- NULL
}
```

# Differential Expression with DESeq2

## Preparing the DESeqDataSet

**Theory: DESeqDataSet construction**

DESeq2 requires raw counts (not normalized), a sample metadata data
frame, and a design formula. We extract the raw counts from the Seurat
object's `counts` layer and the metadata from `meta.data`. The design
formula `~ response2` tells DESeq2 to model gene expression as a
function of response status.

Important: DESeq2 works on raw counts because it applies its own
normalization internally (size factors based on the median-of-ratios
method). We should NOT feed it the Seurat-normalized data.

If SVA detected surrogate variables, we add them to the metadata and
include them in the design formula to adjust for hidden confounders.

```{r prepare-deseq2}
# Extract raw counts from Seurat
raw_counts <- GetAssayData(seurat_obj, layer = "counts")

# Get metadata
meta <- seurat_obj@meta.data

# Ensure response2 is a factor with NonResponder as reference
meta$response2 <- factor(meta$response2,
                          levels = c("NonResponder", "Responder"))

# Add surrogate variables to metadata if detected
if (!is.null(sv_df)) {
  meta <- cbind(meta, sv_df[rownames(meta), , drop = FALSE])
  design_formula <- as.formula(
    paste("~ response2 +", paste(colnames(sv_df), collapse = " + "))
  )
  cat("Design includes response2 +", ncol(sv_df), "surrogate variables\n")
} else {
  design_formula <- ~ response2
  cat("Design includes response2 only (no SVs detected)\n")
}

# Create DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(raw_counts),
  colData   = meta,
  design    = design_formula
)

cat("DESeqDataSet created:", nrow(dds), "genes x", ncol(dds), "samples\n")
```

## Pre-filtering

We apply a light pre-filter to remove genes with very low counts. This
reduces memory usage and speeds up computation. DESeq2's internal
independent filtering will do additional filtering, but pre-filtering
removes obvious noise.

```{r prefilter}
# Keep genes with at least 10 total counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
cat("Genes after pre-filtering:", nrow(dds), "\n")
```

## Running DESeq2

**Theory: What happens inside DESeq()?**

The single `DESeq()` call performs three steps internally:

1. **Estimation of size factors** (`estimateSizeFactors()`): DESeq2
   computes the geometric mean of each gene across all samples, then
   for each sample calculates the median ratio of observed counts to
   these geometric means. This ratio is the size factor.

2. **Estimation of gene-wise dispersions** (`estimateDispersions()`):
   For each gene, DESeq2 estimates the dispersion (a measure of how
   much the variance exceeds the mean). It then fits a smooth curve
   relating mean expression to dispersion and shrinks each gene's
   dispersion toward this trend using an empirical Bayes approach.

3. **Negative binomial GLM fitting and Wald test**
   (`nbinomWaldTest()`): For each gene, DESeq2 fits a generalized
   linear model with the NB distribution and tests whether the log2
   fold change for response2 is significantly different from zero.

```{r run-deseq2}
dds <- DESeq(dds)

cat("DESeq2 analysis complete.\n")
cat("Results names:", resultsNames(dds), "\n")
```

```{r dispersion-plot, fig.cap="Dispersion estimates vs. mean normalized counts. Black dots: gene-wise estimates. Red curve: fitted trend. Blue dots: shrunken (final) estimates."}
plotDispEsts(dds)
```

## Extracting Results

```{r extract-results}
# Extract results for Responder vs NonResponder
res <- results(dds, contrast = c("response2", "Responder", "NonResponder"),
               alpha = 0.05)

cat("DESeq2 results summary:\n")
summary(res, alpha = 0.05)
```

## Log Fold Change Shrinkage

**Theory: Why shrink the LFC?**

Raw LFC estimates from DESeq2 can be unreliable for genes with low
counts or high dispersion. These genes may show extreme fold changes
simply due to noise. LFC shrinkage uses a Bayesian approach to moderate
these estimates: genes with weak evidence (low counts, high variance)
are pulled toward zero, while genes with strong evidence keep their
original LFC.

The `apeglm` method (Zhu, Ibrahim, Love, 2019) is the recommended
shrinkage estimator. It produces adaptive shrinkage that preserves
large, well-supported fold changes while dampening noisy ones.

Note: This step is unique to DESeq2. In Pipeline 1, limma's eBayes
moderation serves a similar purpose for the variance, but does not
shrink the fold changes themselves.

```{r lfc-shrinkage}
resLFC <- lfcShrink(dds,
                     coef = "response2_Responder_vs_NonResponder",
                     type = "apeglm")

cat("After LFC shrinkage:\n")
summary(resLFC, alpha = 0.05)
```

## Comparing Shrunken vs. Unshrunken LFC

```{r ma-comparison, fig.cap="MA plots comparing unshrunken (left) vs. apeglm-shrunken (right) log2 fold changes. Shrinkage reduces noisy extreme values for lowly expressed genes.", fig.width=12, fig.height=5}
par(mfrow = c(1, 2))
plotMA(res, main = "Unshrunken LFC", ylim = c(-5, 5))
plotMA(resLFC, main = "Shrunken LFC (apeglm)", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

# Visualization of DE Results

## Define Strict DEGs

```{r define-strict-degs}
strict_fdr <- 0.05
min_abs_lfc <- 1.0

# Convert DESeq2 results to data frame
res_df <- as.data.frame(resLFC) %>%
  rownames_to_column(var = "gene_id") %>%
  filter(!is.na(padj))

# Merge with gene metadata for symbols
# gene_meta has columns like gene_id, gene_biotype, symbol.
# We use the rownames of gene_meta as the merge key to avoid
# the "Column name must not be duplicated" error.
gene_meta_df <- gene_meta
gene_meta_df$merge_key <- rownames(gene_meta_df)

# Only keep columns not already in res_df (avoid duplicating baseMean etc.)
cols_to_add <- setdiff(colnames(gene_meta_df), colnames(res_df))
cols_to_add <- c("merge_key", cols_to_add)

res_df <- left_join(res_df,
                    gene_meta_df[, cols_to_add, drop = FALSE],
                    by = c("gene_id" = "merge_key"))

# Detect symbol column
sym_col <- intersect(c("symbol", "Symbol", "gene_name"), colnames(res_df))[1]

# Apply strict thresholds
strict_DEGs <- res_df %>%
  filter(padj < strict_fdr, abs(log2FoldChange) >= min_abs_lfc)

cat("Strict DEGs (FDR <", strict_fdr, ", |log2FC| >=", min_abs_lfc, "):",
    nrow(strict_DEGs), "\n")
cat("  Upregulated in Responders:  ", sum(strict_DEGs$log2FoldChange > 0), "\n")
cat("  Downregulated in Responders:", sum(strict_DEGs$log2FoldChange < 0), "\n")
```

## P-value Distribution

**Theory: Interpreting p-value histograms**

A well-calibrated DE analysis should produce a p-value distribution that
is a mixture of a uniform component (from true null genes) and a spike
near zero (from truly DE genes). An anti-conservative pattern (spike
near 1) may indicate model misspecification.

```{r pvalue-hist, fig.cap="Distribution of raw p-values from DESeq2. A spike near zero indicates true DE genes above the uniform null distribution."}
hist(res_df$pvalue, breaks = 50, col = "lightblue", border = "white",
     main = "", xlab = "Raw p-value", ylab = "Number of genes", las = 1)
abline(v = 0.05, col = "blue", lwd = 2, lty = 2)
text(0.3, max(hist(res_df$pvalue, plot = FALSE)$counts) * 0.9,
     paste0("Genes with p <= 0.05: ", sum(res_df$pvalue <= 0.05)),
     col = "blue", font = 2, cex = 0.9)
```

## Volcano Plot

```{r volcano-plot, fig.cap="Volcano plot of shrunken log2 fold changes vs. -log10 adjusted p-values. Red points are strict DEGs. Top genes are labeled.", fig.width=10, fig.height=7}
res_df$is_strict <- res_df$padj < strict_fdr &
                    abs(res_df$log2FoldChange) >= min_abs_lfc &
                    !is.na(res_df$padj)

# Top genes to label
top_to_label <- res_df %>%
  filter(is_strict) %>%
  arrange(padj) %>%
  slice_head(n = 15)

p <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = is_strict), size = 0.8, alpha = 0.6) +
  scale_color_manual(values = c("grey60", "firebrick"),
                     labels = c("Not significant", "Strict DEG")) +
  geom_vline(xintercept = c(-min_abs_lfc, min_abs_lfc),
             linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  geom_hline(yintercept = -log10(strict_fdr),
             linetype = "dashed", color = "blue", linewidth = 0.5) +
  labs(x = expression(log[2] ~ "FC (Responder vs NonResponder)"),
       y = expression(-log[10] ~ "(adjusted p-value)"),
       color = "") +
  theme_minimal() +
  theme(legend.position = "top")

# Add labels if we have symbols
if (!is.na(sym_col) && sym_col %in% colnames(top_to_label)) {
  p <- p + geom_text_repel(data = top_to_label,
                           aes(label = .data[[sym_col]]),
                           size = 3, max.overlaps = 20)
}
print(p)
```

## Heatmap of Top DEGs

```{r heatmap, fig.cap="Heatmap of the top 50 DEGs (by adjusted p-value) using variance-stabilized expression values. Rows are z-scored across samples.", fig.height=10}
# Variance stabilizing transformation for visualization
vsd <- vst(dds, blind = FALSE)

# Top 50 genes
top50_genes <- res_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  slice_head(n = 50) %>%
  pull(gene_id)

# Extract and z-score
mat <- assay(vsd)[top50_genes, ]
mat_z <- t(scale(t(mat)))

# Row labels: use symbols if available
if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
  sym_lookup <- setNames(res_df[[sym_col]], res_df$gene_id)
  row_labels <- ifelse(is.na(sym_lookup[rownames(mat_z)]) |
                        sym_lookup[rownames(mat_z)] == "",
                        rownames(mat_z),
                        sym_lookup[rownames(mat_z)])
  rownames(mat_z) <- row_labels
}

# Annotation
annot_col <- data.frame(
  Response = colData(dds)$response2,
  row.names = colnames(mat_z)
)
annot_colors <- list(
  Response = c(NonResponder = "steelblue", Responder = "coral")
)

pheatmap(mat_z,
         annotation_col = annot_col,
         annotation_colors = annot_colors,
         cluster_rows = TRUE, cluster_cols = TRUE,
         show_colnames = FALSE,
         fontsize_row = 7,
         color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         main = "Top 50 DEGs (z-scored VST values)")
```

## PCA on VST-Transformed Data

**Theory: VST vs. voom**

DESeq2 provides two transformations for visualization: `vst()`
(variance-stabilizing transformation) and `rlog()` (regularized log).
Both stabilize the variance across the range of mean values, making the
data suitable for PCA, clustering, and heatmaps. VST is faster and
recommended for datasets with more than ~30 samples, while rlog is
better for very small datasets (<20 samples). Both are conceptually
similar to what voom does in Pipeline 1, but operate differently:

- **voom** (Pipeline 1): estimates precision weights for each
  observation, keeping the data as weighted log-CPM
- **VST** (Pipeline 2): transforms the counts so that the variance
  becomes approximately constant across expression levels

```{r pca-vst, fig.cap="PCA using VST-transformed data from DESeq2. This is the DESeq2 equivalent of the MDS plot in Pipeline 1."}
plotPCA(vsd, intgroup = "response2") +
  scale_color_manual(values = c("steelblue", "coral")) +
  theme_minimal() +
  ggtitle("PCA on VST-transformed data")
```

## Top 20 DEGs Table

```{r top20-table}
top20 <- strict_DEGs %>%
  arrange(desc(abs(log2FoldChange))) %>%
  slice_head(n = 20)

# Build display columns dynamically
display_cols <- c("gene_id", "log2FoldChange", "baseMean",
                  "stat", "pvalue", "padj")
if (!is.na(sym_col)) display_cols <- c(display_cols, sym_col)

# Only select columns that actually exist
display_cols <- intersect(display_cols, colnames(top20))

top20_display <- top20 %>%
  dplyr::select(all_of(display_cols))

kable(top20_display,
      caption = "Top 20 DEGs by absolute log2 fold change (DESeq2, FDR < 0.05, |log2FC| >= 1).",
      digits = 4,
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Gene-Level Boxplots (Top 6)

```{r gene-boxplots, fig.cap="Normalized counts for the top 6 DEGs, comparing Responders vs NonResponders.", fig.width=12, fig.height=8}
top6_genes <- strict_DEGs %>%
  arrange(padj) %>%
  slice_head(n = 6) %>%
  pull(gene_id)

par(mfrow = c(2, 3))
for (gene in top6_genes) {
  # Get symbol if available
  label <- gene
  if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
    s <- res_df[[sym_col]][res_df$gene_id == gene]
    if (length(s) > 0 && !is.na(s[1]) && s[1] != "") label <- s[1]
  }
  plotCounts(dds, gene = gene, intgroup = "response2",
             main = label, col = c("steelblue", "coral"))
}
par(mfrow = c(1, 1))
```

## Save DE Results

```{r save-results}
# Full results
write.csv(res_df, "../results/tables/p2_full_DE_results_deseq2.csv",
          row.names = FALSE)

# Strict DEGs
write.csv(strict_DEGs, "../results/tables/p2_strict_DEGs_deseq2.csv",
          row.names = FALSE)

cat("Saved DESeq2 results to results/tables/\n")
```

# Functional Enrichment Analysis

## GO Enrichment with GOstats

**Theory: Hypergeometric test for GO enrichment**

Gene Ontology (GO) enrichment asks: among our list of DE genes, are
certain biological processes over-represented compared to what we would
expect by chance? The hypergeometric test models this as drawing balls
from an urn: given N total genes, K genes annotated to a GO term, and n
DE genes, how many DE genes annotated to that term (k) would we expect?
A significant p-value means the overlap is larger than chance.

The conditional test accounts for the hierarchical structure of GO by
conditioning on child terms that have already been found significant,
reducing redundancy.

```{r go-enrichment}
# Map DEG Ensembl IDs to Entrez IDs
deg_genes <- strict_DEGs$gene_id

entrez_map <- tryCatch({
  AnnotationDbi::select(org.Hs.eg.db,
                        keys = deg_genes,
                        columns = "ENTREZID",
                        keytype = "ENSEMBL")
}, error = function(e) {
  # If Ensembl fails, try with gene symbols
  if (!is.na(sym_col) && sym_col %in% colnames(strict_DEGs)) {
    syms <- strict_DEGs[[sym_col]]
    syms <- syms[!is.na(syms) & syms != ""]
    AnnotationDbi::select(org.Hs.eg.db,
                          keys = syms,
                          columns = "ENTREZID",
                          keytype = "SYMBOL")
  } else {
    data.frame(ENTREZID = character(0))
  }
})

entrez_ids <- unique(na.omit(entrez_map$ENTREZID))
cat("Mapped", length(entrez_ids), "unique Entrez IDs from",
    length(deg_genes), "DEGs\n")
```

```{r run-go-enrichment}
if (length(entrez_ids) > 5) {
  params <- new("GOHyperGParams",
                geneIds      = entrez_ids,
                annotation   = "org.Hs.eg.db",
                ontology     = "BP",
                pvalueCutoff = 0.05,
                testDirection = "over")
  conditional(params) <- TRUE
  hgOverCond <- hyperGTest(params)

  # Summarize and filter
  resGO <- summary(hgOverCond)
  resGO <- resGO[resGO$Size >= 3 & resGO$Size <= 300 & resGO$Count >= 3, ]
  resGO <- resGO[order(resGO$OddsRatio, decreasing = TRUE), ]

  cat("Significant GO BP terms after filtering:", nrow(resGO), "\n")
} else {
  cat("Too few mapped genes for enrichment analysis.\n")
  resGO <- NULL
}
```

```{r go-pvalue-dist, fig.cap="P-value distribution for GO enrichment before and after filtering.", fig.width=10, fig.height=4}
if (!is.null(resGO) && nrow(resGO) > 0) {
  pvals_all <- pvalues(hgOverCond)
  par(mfrow = c(1, 2), mar = c(4, 5, 2, 1))
  hist(pvals_all, xlab = "P-value", main = "Before filtering", las = 1,
       col = "lightblue", border = "white")
  hist(pvals_all[resGO$GOBPID], xlab = "P-value", main = "After filtering",
       las = 1, col = "lightblue", border = "white", xlim = c(0, 1))
  par(mfrow = c(1, 1))
}
```

```{r go-gene-symbols}
if (!is.null(resGO) && nrow(resGO) > 0) {
  # Map Entrez IDs to gene symbols for each GO term
  geneIDs_by_GO <- geneIdsByCategory(hgOverCond)[resGO$GOBPID]
  all_entrez <- unique(as.character(unlist(geneIDs_by_GO)))

  entrez_to_symbol <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = all_entrez,
    columns = "SYMBOL",
    keytype = "ENTREZID"
  )
  sym_map_go <- setNames(entrez_to_symbol$SYMBOL, entrez_to_symbol$ENTREZID)

  gene_symbols_by_go <- sapply(geneIDs_by_GO, function(ids) {
    syms <- sym_map_go[as.character(ids)]
    paste(syms[!is.na(syms)], collapse = ", ")
  })

  go_results <- cbind(resGO, Genes = unlist(gene_symbols_by_go))
  rownames(go_results) <- seq_len(nrow(go_results))
}
```

```{r go-table}
if (!is.null(resGO) && nrow(resGO) > 0) {
  kable(head(go_results, 15),
        caption = "Top 15 enriched GO Biological Process terms.",
        row.names = FALSE) %>%
    kable_styling(full_width = FALSE, position = "center",
                  bootstrap_options = c("striped", "hover", "condensed")) %>%
    scroll_box(width = "100%", height = "400px")
}
```

## GSEA with fgsea

**Theory: Ranked-list GSEA**

GSEA uses the full ranked gene list rather than a binary DEG list. We
rank genes by the DESeq2 Wald statistic (which combines fold change
magnitude and significance) and test for pathway enrichment.

```{r gsea-ranked-list}
# # Create ranked list using DESeq2 stat (Wald statistic)
# if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
#   all_symbols <- res_df[[sym_col]]
#   all_stats   <- res_df$stat
# 
#   # Remove NAs
#   valid <- !is.na(all_symbols) & all_symbols != "" & !is.na(all_stats)
#   gene_list <- setNames(all_stats[valid], all_symbols[valid])
# 
#   # Remove duplicates (keep highest absolute stat)
#   gene_list <- gene_list[order(abs(gene_list), decreasing = TRUE)]
#   gene_list <- gene_list[!duplicated(names(gene_list))]
# 
#   # Sort decreasingly
#   gene_list <- sort(gene_list, decreasing = TRUE)
# 
#   cat("GSEA ranked list:", length(gene_list), "genes\n")
#   cat("Top 5:", head(names(gene_list)), "\n")
#   cat("Bottom 5:", tail(names(gene_list)), "\n")
# }
```

```{r save-enrichment}
if (exists("go_results") && nrow(go_results) > 0) {
  write.csv(go_results, "../results/enrichment/p2_GO_BP_results.csv",
            row.names = FALSE)
  cat("Saved enrichment results to results/enrichment/\n")
}
```

# Comparison with Pipeline 1

## Summary Table

```{r comparison}
cat("=== Pipeline Comparison ===\n\n")
cat("Pipeline 1 (edgeR/limma-voom):\n")
cat("  Container:      SummarizedExperiment + DGEList\n")
cat("  Normalization:  TMM\n")
cat("  Transformation: voom (precision weights)\n")
cat("  DE test:        Moderated t-test (eBayes)\n")
cat("  Enrichment:     GOstats (hypergeometric)\n\n")

cat("Pipeline 2 (Seurat + DESeq2):\n")
cat("  Container:      Seurat object\n")
cat("  Normalization:  LogNormalize (Seurat) + median-of-ratios (DESeq2)\n")
cat("  Distribution:   Negative binomial (counts directly)\n")
cat("  DE test:        Wald test + apeglm LFC shrinkage\n")
cat("  Enrichment:     GOstats (hypergeometric)\n\n")

cat("Strict DEGs found (Pipeline 2):", nrow(strict_DEGs), "\n")
```

# Discussion

In this second pipeline, we analyzed the same CLL dasatinib response
dataset using the Seurat + DESeq2 framework. The key differences from
Pipeline 1 include:

1. **Data container**: The Seurat object provides a unified interface
   for normalization, scaling, dimensionality reduction, and
   visualization. Unlike the SummarizedExperiment + DGEList combination,
   Seurat handles all steps in a single object with consistent syntax.

2. **Statistical model**: DESeq2's negative binomial model fits raw
   counts directly, avoiding the need for voom transformation. The
   dispersion shrinkage (visible in the dispersion plot) serves a
   similar purpose to limma's eBayes variance moderation.

3. **LFC shrinkage**: The apeglm shrinkage method in DESeq2 moderates
   fold change estimates, producing more reliable effect sizes for
   lowly expressed genes. This is unique to the DESeq2 framework and
   does not have a direct equivalent in limma-voom.

4. **UMAP visualization**: Unlike Pipeline 1 which uses MDS, we
   applied both PCA and UMAP here via Seurat. For our small sample
   size (28 samples), we adapted the n.neighbors parameter to avoid
   the "n_neighbors must be smaller than dataset size" error.

5. **SVA integration**: We included Surrogate Variable Analysis to
   detect and correct for hidden confounders, with the surrogate
   variables incorporated directly into the DESeq2 design formula.

The biological conclusions (enrichment of apoptosis and B-cell signaling
pathways among DEGs in responders) should be consistent across both
pipelines, validating the robustness of our findings regardless of the
analytical framework.

# Session Information

```{r session-info}
sessionInfo()
```

# References
