---
title: "Bulk RNA-seq Analysis of CLL Dasatinib Response"
subtitle: "Seurat Object + DESeq2 Approach"
author:
- name: "Samuel Escudero, Ivon Sanchez, Karim Hamed"
  affiliation: Universitat Pompeu Fabra
  email: samuel.escudero01@estudiant.upf.edu
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_captions: yes
    code_folding: show
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "",
  fig.align   = "center",
  fig.width   = 8,
  fig.height  = 6,
  fig.wide    = TRUE,
  cache       = FALSE,
  message     = FALSE,
  warning     = FALSE
)
```

# Introduction

## Biological context

This is **Pipeline 2** of our multi-approach analysis of the GSE151159
CLL dataset. We are analyzing the same data as Pipeline 1 (edgeR/limma-
voom), but now using a **Seurat object** as our data container and
**DESeq2** as our differential expression framework. By running
parallel pipelines on identical data, we can directly compare how
different data structures, normalization strategies, and statistical
models affect our biological conclusions.

For the full biological context (CLL, dasatinib, responders vs.
non-responders), refer to Pipeline 1. Here we focus on the
**methodological differences** between the two approaches.

## Why Seurat + DESeq2?

**Theory: The Seurat object**

Seurat was originally developed for single-cell RNA-seq (scRNA-seq)
analysis by the Satija Lab (Butler et al., 2018; Stuart et al., 2019).
However, its data structure is perfectly capable of handling bulk
RNA-seq data. A Seurat object stores:

- **Assays**: named slots containing expression matrices. The default
  `RNA` assay holds raw counts in `counts`, normalized data in `data`,
  and scaled data in `scale.data`.
- **Meta.data**: a data frame of sample-level (or cell-level in scRNA-seq)
  metadata, equivalent to `colData()` in SummarizedExperiment.
- **Reductions**: dimensionality reduction results (PCA, UMAP, tSNE).
- **Graphs**: nearest-neighbor graphs for clustering.

The advantage of using Seurat for bulk RNA-seq is that it provides a
unified interface for normalization, dimensionality reduction, and
visualization, with built-in functions like `NormalizeData()`,
`ScaleData()`, `RunPCA()`, and `RunUMAP()`. It also makes it easy to
transition between bulk and single-cell workflows.

**Theory: DESeq2 vs. limma-voom**

While Pipeline 1 used `voom()` to transform counts into weighted
log-CPM values and then applied linear models, DESeq2 takes a
fundamentally different approach:

1. **Negative binomial model**: DESeq2 models the raw counts directly
   using a negative binomial (NB) distribution, which naturally handles
   the overdispersion (variance > mean) typical of RNA-seq data. No
   transformation is needed.

2. **Shrinkage of dispersion estimates**: DESeq2 estimates gene-wise
   dispersions and then shrinks them toward a fitted trend using an
   empirical Bayes procedure. This is conceptually similar to what
   limma's `eBayes()` does for variances, but operates on the
   dispersion parameter of the NB distribution.

3. **Wald test**: For each gene, DESeq2 tests whether the log2 fold
   change coefficient is significantly different from zero using a
   Wald test (ratio of the estimated LFC to its standard error).

4. **Log fold change shrinkage**: After testing, DESeq2 can apply
   LFC shrinkage using methods like `apeglm`, which produces more
   reliable effect size estimates for genes with low counts or high
   dispersion. This is unique to DESeq2 and particularly useful for
   ranking genes and visualization.

The key practical differences are:

| Feature | limma-voom (Pipeline 1) | DESeq2 (Pipeline 2) |
|---------|------------------------|---------------------|
| Input | Transformed log-CPM + weights | Raw counts directly |
| Distribution | Normal (after voom) | Negative binomial |
| Variance stabilization | voom precision weights | NB dispersion shrinkage |
| Test | Moderated t-test | Wald test |
| LFC shrinkage | Not built-in | apeglm/ashr/normal |
| Speed | Faster | Slower (iterative fitting) |
| Best for | Large samples, complex designs | Small-moderate samples |

## Libraries

```{r load-libraries}
# Core packages
library(Seurat)               # Data container and visualization
library(DESeq2)               # Differential expression (NB model)
library(SummarizedExperiment)  # To read the RDS initially

# Visualization
library(ggplot2)              # Grammar of graphics
library(ggrepel)              # Non-overlapping labels
library(pheatmap)             # Clustered heatmaps
library(RColorBrewer)         # Color palettes

# Functional enrichment
#library(clusterProfiler)      # ORA and GSEA
library(org.Hs.eg.db)         # Human gene annotations
library(AnnotationDbi)        # Annotation interface
#library(enrichplot)           # Enrichment visualization

# Data manipulation and reporting
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
```

# Data Import and Seurat Object Construction

## Loading the data

We load the shared CSV files exported by Pipeline 1. This ensures
both pipelines analyze exactly the same set of samples and genes.

If you have not run Pipeline 1's export step, you can also load
directly from the RDS and export here.

```{r load-data}
# Option A: Load from shared CSVs (preferred, ensures consistency)
if (file.exists("../data/counts_matrix.csv")) {
  counts_mat <- read.csv("../data/counts_matrix.csv", row.names = 1,
                         check.names = FALSE)
  sample_meta <- read.csv("../data/sample_metadata.csv", row.names = 1,
                          check.names = FALSE)
  gene_meta <- read.csv("../data/gene_metadata.csv", row.names = 1,
                        check.names = FALSE)
  cat("Loaded from shared CSVs\n")

# Option B: Load from RDS directly
} else {
  cat("CSVs not found. Loading from RDS and exporting...\n")
  cll <- readRDS("../data/GSE151159.rds")

  # Create response classification
  desc2 <- colData(cll)$description.2
  response_raw <- ifelse(
    grepl("^responders_final", desc2), "Responder",
    ifelse(grepl("^nonresponders_final", desc2), "NonResponder", NA)
  )
  colData(cll)$response2 <- factor(response_raw,
                                    levels = c("NonResponder", "Responder"))

  # Remove ambiguous samples
  keep <- !is.na(colData(cll)$response2)
  cll <- cll[, keep]

  counts_mat <- as.data.frame(assays(cll)$counts)
  sample_meta <- as.data.frame(colData(cll))
  gene_meta <- as.data.frame(rowData(cll))

  # Export for future use
  write.csv(counts_mat, "../data/counts_matrix.csv", row.names = TRUE)
  write.csv(sample_meta, "../data/sample_metadata.csv", row.names = TRUE)
  write.csv(gene_meta, "../data/gene_metadata.csv", row.names = TRUE)
}

cat("Count matrix:", nrow(counts_mat), "genes x", ncol(counts_mat), "samples\n")
cat("Sample groups:\n")
table(sample_meta$response2)
```

## Creating the Seurat Object

**Theory: CreateSeuratObject internals**

When we call `CreateSeuratObject()`, Seurat performs several things
under the hood:

1. Creates an `Assay` object (v5) or `Assay5` containing the count
   matrix in the `counts` layer
2. Computes `nCount_RNA` (total counts per sample, equivalent to
   library size) and `nFeature_RNA` (number of detected genes per
   sample)
3. Stores sample metadata in the `meta.data` slot
4. Sets the default assay to `RNA`

The `min.cells` and `min.features` parameters are filtering thresholds.
For bulk RNA-seq, `min.cells` controls how many samples must express a
gene for it to be retained. We set `min.cells = 3` as a light
pre-filter (genes must be detected in at least 3 samples).

```{r create-seurat}
# Ensure response2 is a factor
sample_meta$response2 <- factor(sample_meta$response2,
                                 levels = c("NonResponder", "Responder"))

# Create Seurat object
seurat_obj <- CreateSeuratObject(
  counts    = as.matrix(counts_mat),
  meta.data = sample_meta,
  project   = "CLL_Dasatinib",
  min.cells = 3,       # Gene must be in >= 3 samples
  min.features = 200   # Sample must have >= 200 detected genes
)

cat("Seurat object created:\n")
seurat_obj
```

```{r inspect-seurat}
# Inspect the structure
cat("Default assay:", DefaultAssay(seurat_obj), "\n")
cat("Assay layers:", Layers(seurat_obj), "\n")
cat("Metadata columns:", colnames(seurat_obj@meta.data), "\n\n")

# Check nCount and nFeature distributions
cat("Library size (nCount_RNA) range:",
    range(seurat_obj$nCount_RNA), "\n")
cat("Detected genes (nFeature_RNA) range:",
    range(seurat_obj$nFeature_RNA), "\n")
```

# Quality Control

## QC Metrics Visualization

In scRNA-seq, we would filter cells based on mitochondrial percentage,
doublet scores, and gene/UMI counts. For bulk RNA-seq, these metrics
are less critical because each "cell" is actually a bulk sample with
millions of reads. However, visualizing library size and gene detection
rates helps us identify potential outlier samples.

```{r qc-violin, fig.cap="Distribution of QC metrics across samples. Left: total read counts (library size). Right: number of detected genes per sample."}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA"),
        group.by = "response2",
        pt.size = 0.5,
        cols = c("steelblue", "coral")) +
  theme(axis.text.x = element_text(angle = 0))
```

```{r qc-scatter, fig.cap="Relationship between library size and number of detected genes, colored by response group. A strong linear relationship is expected for good-quality samples."}
FeatureScatter(seurat_obj,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA",
               group.by = "response2",
               cols = c("steelblue", "coral"))
```

# Normalization

## LogNormalize (Seurat default)

**Theory: Seurat's LogNormalize**

Seurat's default normalization uses a simple approach:

1. Divide each gene's count by the total counts for that sample
   (equivalent to calculating a proportion)
2. Multiply by a scale factor (default: 10,000)
3. Take the natural log (log1p, i.e., log(x + 1))

This is conceptually similar to computing log-CPM (counts per million),
but with a scale factor of 10,000 instead of 1,000,000. The formula is:

`normalized_value = log(count / total_counts * scale_factor + 1)`

This is simpler than TMM (which adjusts for RNA composition) but works
well in practice for many datasets. The normalized values are stored in
the `data` layer of the RNA assay.

```{r normalize}
seurat_obj <- NormalizeData(
  seurat_obj,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)

cat("Normalization complete.\n")
cat("Layers after normalization:", Layers(seurat_obj), "\n")
```

## Identify Highly Variable Features

**Theory: Why select variable features?**

For dimensionality reduction (PCA, UMAP), we want to focus on genes
that show meaningful variation across samples, not genes with constant
expression. Highly variable genes (HVGs) are those with the most
variance relative to their mean expression. Seurat uses the `vst`
(variance-stabilizing transformation) method to identify HVGs,
which models the relationship between mean and variance and selects
genes that deviate most from this trend.

In scRNA-seq, we typically select 2000-3000 HVGs. For bulk RNA-seq with
fewer samples, we can be more generous (3000-5000) to capture more
of the biological variation.

```{r find-variable-features, fig.cap="Highly variable genes identified by the VST method. Red points are the selected variable features. Top 10 are labeled."}
seurat_obj <- FindVariableFeatures(
  seurat_obj,
  selection.method = "vst",
  nfeatures = 3000
)

# Plot variable features
top10_hvg <- head(VariableFeatures(seurat_obj), 10)
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10_hvg, repel = TRUE)
plot2
```

## Scale Data

**Theory: ScaleData**

Scaling performs two operations per gene:

1. **Centers** the expression values (subtracts the mean across samples)
2. **Scales** by dividing by the standard deviation

After scaling, each gene has mean = 0 and SD = 1. This is essential for
PCA because without scaling, PCA would be dominated by highly expressed
genes simply because they have larger absolute values, not because they
carry more biological information. The scaled data is stored in the
`scale.data` layer.

```{r scale-data}
seurat_obj <- ScaleData(seurat_obj, features = rownames(seurat_obj))
cat("Data scaled. Layers:", Layers(seurat_obj), "\n")
```

# Dimensionality Reduction

## PCA

**Theory: PCA on bulk RNA-seq via Seurat**

PCA (Principal Component Analysis) finds the axes of maximum variance
in the data. Seurat's `RunPCA()` operates on the scaled data and
computes the top principal components. For bulk RNA-seq, PC1 and PC2
typically capture the dominant biological signal (e.g., treatment
response), while later PCs may capture batch effects or noise.

The **elbow plot** helps us decide how many PCs contain real signal:
we look for the "elbow" where the standard deviation drops sharply,
indicating that subsequent PCs capture mostly noise.

Note that in Pipeline 1 we used MDS (multidimensional scaling) with
leading log-fold change distances. PCA and MDS are conceptually
similar but not identical: MDS preserves sample distances directly,
while PCA finds the directions of maximum variance. For well-normalized
bulk RNA-seq, both should give similar sample groupings.

```{r run-pca}
seurat_obj <- RunPCA(seurat_obj,
                      features = VariableFeatures(seurat_obj),
                      npcs = 20)
```

```{r pca-plot, fig.cap="PCA of samples colored by dasatinib response status. Each point is one sample."}
DimPlot(seurat_obj, reduction = "pca", group.by = "response2",
        cols = c("steelblue", "coral"), pt.size = 3) +
  ggtitle("PCA: Responders vs NonResponders")
```

```{r elbow-plot, fig.cap="Elbow plot showing the standard deviation explained by each principal component. The elbow point suggests how many PCs carry meaningful signal."}
ElbowPlot(seurat_obj, ndims = 20)
```

```{r pca-loadings, fig.cap="Top genes contributing to PC1 and PC2. These genes drive the primary separation between samples."}
VizDimLoadings(seurat_obj, dims = 1:2, reduction = "pca")
```

## UMAP

**Theory: UMAP vs. PCA**

UMAP (Uniform Manifold Approximation and Projection) is a nonlinear
dimensionality reduction technique that preserves both local and global
structure better than PCA. While PCA is limited to linear relationships,
UMAP can capture complex, nonlinear patterns in the data. However, UMAP
is stochastic (results vary between runs) and distances in UMAP space
are not directly interpretable. For bulk RNA-seq with few samples
(~28), UMAP adds limited value over PCA, but it's standard practice in
Seurat workflows and provides a nice visualization.

```{r run-umap, fig.cap="UMAP projection of samples colored by response status."}
seurat_obj <- RunUMAP(seurat_obj, dims = 1:15)

DimPlot(seurat_obj, reduction = "umap", group.by = "response2",
        cols = c("steelblue", "coral"), pt.size = 2) +
  ggtitle("UMAP: Responders vs NonResponders")
```

# Differential Expression with DESeq2

## Preparing the DESeqDataSet

**Theory: DESeqDataSet construction**

DESeq2 requires raw counts (not normalized), a sample metadata data
frame, and a design formula. We extract the raw counts from the Seurat
object's `counts` layer and the metadata from `meta.data`. The design
formula `~ response2` tells DESeq2 to model gene expression as a
function of response status.

Important: DESeq2 works on raw counts because it applies its own
normalization internally (size factors based on the median-of-ratios
method). We should NOT feed it the Seurat-normalized data.

```{r prepare-deseq2}
# Extract raw counts from Seurat
raw_counts <- GetAssayData(seurat_obj, layer = "counts")

# Get metadata
meta <- seurat_obj@meta.data

# Ensure response2 is a factor with NonResponder as reference
meta$response2 <- factor(meta$response2,
                          levels = c("NonResponder", "Responder"))

# Create DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(raw_counts),
  colData   = meta,
  design    = ~ response2
)

cat("DESeqDataSet created:", nrow(dds), "genes x", ncol(dds), "samples\n")
```

## Pre-filtering

We apply a light pre-filter to remove genes with very low counts. This
reduces memory usage and speeds up computation. DESeq2's internal
independent filtering will do additional filtering, but pre-filtering
removes obvious noise.

```{r prefilter}
# Keep genes with at least 10 total counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
cat("Genes after pre-filtering:", nrow(dds), "\n")
```

## Running DESeq2

**Theory: What happens inside DESeq()?**

The single `DESeq()` call performs three steps internally:

1. **Estimation of size factors** (`estimateSizeFactors()`): DESeq2
   computes the geometric mean of each gene across all samples, then
   for each sample calculates the median ratio of observed counts to
   these geometric means. This ratio is the size factor. Unlike TMM,
   which uses trimmed M-values, the median-of-ratios method is more
   robust to outlier genes.

2. **Estimation of gene-wise dispersions** (`estimateDispersions()`):
   For each gene, DESeq2 estimates the dispersion (a measure of how
   much the variance exceeds the mean, specific to the NB distribution).
   It then fits a smooth curve relating mean expression to dispersion
   and shrinks each gene's dispersion toward this trend using an
   empirical Bayes approach. The dispersion plot is diagnostic:
   gene-wise estimates (black dots) should scatter around the fitted
   trend (red line), and the shrunken estimates (blue dots) should
   cluster tightly around it.

3. **Negative binomial GLM fitting and Wald test**
   (`nbinomWaldTest()`): For each gene, DESeq2 fits a generalized
   linear model with the NB distribution and tests whether the log2
   fold change for the response2 coefficient is significantly different
   from zero.

```{r run-deseq2}
dds <- DESeq(dds)

cat("DESeq2 analysis complete.\n")
cat("Results names:", resultsNames(dds), "\n")
```

```{r dispersion-plot, fig.cap="Dispersion estimates vs. mean normalized counts. Black dots: gene-wise estimates. Red curve: fitted trend. Blue dots: shrunken (final) estimates. Well-behaved data shows shrunken estimates clustering around the trend."}
plotDispEsts(dds)
```

## Extracting Results

```{r extract-results}
# Extract results for Responder vs NonResponder
res <- results(dds, contrast = c("response2", "Responder", "NonResponder"),
               alpha = 0.05)

cat("DESeq2 results summary:\n")
summary(res, alpha = 0.05)
```

## Log Fold Change Shrinkage

**Theory: Why shrink the LFC?**

Raw LFC estimates from DESeq2 can be unreliable for genes with low
counts or high dispersion. These genes may show extreme fold changes
simply due to noise. LFC shrinkage uses a Bayesian approach to moderate
these estimates: genes with weak evidence (low counts, high variance)
are pulled toward zero, while genes with strong evidence keep their
original LFC.

The `apeglm` method (Zhu, Ibrahim, Love, 2019) is the recommended
shrinkage estimator. It produces adaptive shrinkage that preserves
large, well-supported fold changes while dampening noisy ones. This
makes the shrunken LFCs better for:

- Ranking genes by effect size
- Visualization (volcano plots, MA plots)
- Downstream gene set enrichment analysis

Note: This step is unique to DESeq2. In Pipeline 1, limma's eBayes
moderation serves a similar purpose for the variance, but does not
shrink the fold changes themselves.

```{r lfc-shrinkage}
resLFC <- lfcShrink(dds,
                     coef = "response2_Responder_vs_NonResponder",
                     type = "apeglm")

cat("After LFC shrinkage:\n")
summary(resLFC, alpha = 0.05)
```

## Comparing Shrunken vs. Unshrunken LFC

```{r ma-comparison, fig.cap="MA plots comparing unshrunken (left) vs. apeglm-shrunken (right) log2 fold changes. Shrinkage reduces noisy extreme values for lowly expressed genes.", fig.width=12, fig.height=5}
par(mfrow = c(1, 2))
plotMA(res, main = "Unshrunken LFC", ylim = c(-5, 5))
plotMA(resLFC, main = "Shrunken LFC (apeglm)", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

# Visualization of DE Results

## Define Strict DEGs

```{r define-strict-degs}
strict_fdr <- 0.05
min_abs_lfc <- 1.0

# Convert DESeq2 results to data frame
res_df <- as.data.frame(resLFC) %>%
  rownames_to_column(var = "gene_id") %>%
  filter(!is.na(padj))

# Merge with gene metadata for symbols
# The gene_meta CSV already has a "gene_id" column in its data,
# so we must NOT use rownames_to_column again. Instead we use
# the existing rownames as a merge key.
if (exists("gene_meta")) {
  gene_meta_df <- gene_meta
  gene_meta_df$gene_id_merge <- rownames(gene_meta_df)
  
  # Only keep columns that do not already exist in res_df
  # (avoid duplicating baseMean, log2FoldChange, etc.)
  cols_to_add <- setdiff(colnames(gene_meta_df), colnames(res_df))
  cols_to_add <- c("gene_id_merge", cols_to_add)
  
  res_df <- left_join(res_df,
                      gene_meta_df[, cols_to_add, drop = FALSE],
                      by = c("gene_id" = "gene_id_merge"))
}

# Apply strict thresholds
strict_DEGs <- res_df %>%
  filter(padj < strict_fdr, abs(log2FoldChange) >= min_abs_lfc)

cat("Strict DEGs (FDR <", strict_fdr, ", |log2FC| >=", min_abs_lfc, "):",
    nrow(strict_DEGs), "\n")
cat("  Upregulated in Responders:  ", sum(strict_DEGs$log2FoldChange > 0), "\n")
cat("  Downregulated in Responders:", sum(strict_DEGs$log2FoldChange < 0), "\n")
```

## P-value Distribution

```{r pvalue-hist, fig.cap="Distribution of raw p-values from DESeq2. A spike near zero indicates true DE genes above the uniform null distribution."}
hist(res_df$pvalue, breaks = 50, col = "lightblue", border = "white",
     main = "", xlab = "Raw p-value", ylab = "Number of genes", las = 1)
abline(v = 0.05, col = "blue", lwd = 2, lty = 2)
```

## Volcano Plot

```{r volcano-plot, fig.cap="Volcano plot of shrunken log2 fold changes vs. -log10 adjusted p-values. Red points are strict DEGs. Top genes are labeled.", fig.width=10, fig.height=7}
res_df$is_strict <- res_df$padj < strict_fdr &
                    abs(res_df$log2FoldChange) >= min_abs_lfc &
                    !is.na(res_df$padj)

# Get symbol column name (could be 'symbol' or 'Symbol')
sym_col <- intersect(c("symbol", "Symbol"), colnames(res_df))[1]

# Top genes to label
top_to_label <- res_df %>%
  filter(is_strict) %>%
  arrange(padj) %>%
  slice_head(n = 15)

p <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = is_strict), size = 0.8, alpha = 0.6) +
  scale_color_manual(values = c("grey60", "firebrick"),
                     labels = c("Not significant", "Strict DEG")) +
  geom_vline(xintercept = c(-min_abs_lfc, min_abs_lfc),
             linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  geom_hline(yintercept = -log10(strict_fdr),
             linetype = "dashed", color = "blue", linewidth = 0.5) +
  labs(x = expression(log[2] ~ "FC (Responder vs NonResponder)"),
       y = expression(-log[10] ~ "(adjusted p-value)"),
       color = "") +
  theme_minimal() +
  theme(legend.position = "top")

# Add labels if we have symbols
if (!is.na(sym_col) && sym_col %in% colnames(top_to_label)) {
  p <- p + geom_text_repel(data = top_to_label,
                           aes_string(label = sym_col),
                           size = 3, max.overlaps = 20)
}
print(p)
```

## Heatmap of Top DEGs

```{r heatmap, fig.cap="Heatmap of the top 50 DEGs (by adjusted p-value) using variance-stabilized expression values. Rows are z-scored across samples.", fig.height=10}
# Variance stabilizing transformation for visualization
vsd <- vst(dds, blind = FALSE)

# Top 50 genes
top50_genes <- res_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  slice_head(n = 50) %>%
  pull(gene_id)

# Extract and z-score
mat <- assay(vsd)[top50_genes, ]
mat_z <- t(scale(t(mat)))

# Row labels: use symbols if available
if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
  sym_lookup <- setNames(res_df[[sym_col]], res_df$gene_id)
  row_labels <- ifelse(is.na(sym_lookup[rownames(mat_z)]) |
                        sym_lookup[rownames(mat_z)] == "",
                        rownames(mat_z),
                        sym_lookup[rownames(mat_z)])
  rownames(mat_z) <- row_labels
}

# Annotation
annot_col <- data.frame(
  Response = colData(dds)$response2,
  row.names = colnames(mat_z)
)
annot_colors <- list(
  Response = c(NonResponder = "steelblue", Responder = "coral")
)

pheatmap(mat_z,
         annotation_col = annot_col,
         annotation_colors = annot_colors,
         cluster_rows = TRUE, cluster_cols = TRUE,
         show_colnames = FALSE,
         fontsize_row = 7,
         color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         main = "Top 50 DEGs (z-scored VST values)")
```

## PCA on VST-Transformed Data

**Theory: VST vs. voom**

DESeq2 provides two transformations for visualization: `vst()`
(variance-stabilizing transformation) and `rlog()` (regularized log).
Both stabilize the variance across the range of mean values, making the
data suitable for PCA, clustering, and heatmaps. VST is faster and
recommended for datasets with more than ~30 samples, while rlog is
better for very small datasets (<20 samples). Both are conceptually
similar to what voom does in Pipeline 1, but operate differently:

- **voom** (Pipeline 1): estimates precision weights for each
  observation, keeping the data as weighted log-CPM
- **VST** (Pipeline 2): transforms the counts so that the variance
  becomes approximately constant across expression levels

```{r pca-vst, fig.cap="PCA using VST-transformed data from DESeq2. This is the DESeq2 equivalent of the MDS plot in Pipeline 1."}
plotPCA(vsd, intgroup = "response2") +
  scale_color_manual(values = c("steelblue", "coral")) +
  theme_minimal() +
  ggtitle("PCA on VST-transformed data")
```

## Top 20 DEGs Table

```{r top20-table}
top20 <- strict_DEGs %>%
  arrange(desc(abs(log2FoldChange))) %>%
  slice_head(n = 20)

# Detect which symbol column exists
sym_col <- intersect(c("symbol", "Symbol", "gene_name"), colnames(top20))[1]

# Build display columns dynamically based on what exists
display_cols <- c("gene_id", "log2FoldChange", "baseMean",
                  "stat", "pvalue", "padj")
if (!is.na(sym_col)) display_cols <- c(display_cols, sym_col)

# Only select columns that actually exist in the data
display_cols <- intersect(display_cols, colnames(top20))

top20_display <- top20 %>%
  dplyr::select(all_of(display_cols))

kable(top20_display,
      caption = "Top 20 DEGs by absolute log2 fold change (DESeq2, FDR < 0.05, |log2FC| >= 1).",
      digits = 4,
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Gene-Level Boxplots (Top 6)

```{r gene-boxplots, fig.cap="Normalized counts for the top 6 DEGs, comparing Responders vs NonResponders.", fig.width=12, fig.height=8}
top6_genes <- strict_DEGs %>%
  arrange(padj) %>%
  slice_head(n = 6) %>%
  pull(gene_id)

par(mfrow = c(2, 3))
for (gene in top6_genes) {
  # Get symbol if available
  label <- gene
  if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
    s <- res_df[[sym_col]][res_df$gene_id == gene]
    if (length(s) > 0 && !is.na(s[1]) && s[1] != "") label <- s[1]
  }
  plotCounts(dds, gene = gene, intgroup = "response2",
             main = label, col = c("steelblue", "coral"))
}
par(mfrow = c(1, 1))
```

## Save DE Results

```{r save-results}
# Full results
#write.csv(res_df, "../results/tables/p2_full_DE_results_deseq2.csv",
#          row.names = FALSE)

# Strict DEGs
#write.csv(strict_DEGs, "../results/tables/p2_strict_DEGs_deseq2.csv",
#          row.names = FALSE)

#cat("Saved DESeq2 results to results/tables/\n")
```

# Functional Enrichment Analysis

## ORA with clusterProfiler

**Theory: clusterProfiler vs. GOstats**

In Pipeline 1, we used `GOstats` for hypergeometric GO testing.
Here, we use `clusterProfiler`, which provides a more modern
interface with built-in visualization functions. Both perform
over-representation analysis (ORA), but clusterProfiler offers
several advantages:

- Supports multiple databases (GO, KEGG, Reactome, MSigDB)
- Built-in visualization (dot plots, enrichment maps, gene-concept
  networks)
- Supports both ORA and GSEA in the same package
- Better handling of gene ID conversion

```{r enrichment-go}
# We need Entrez IDs for clusterProfiler
# First, try to map from Ensembl IDs
deg_genes <- strict_DEGs$gene_id

# Try mapping Ensembl to Entrez
entrez_map <- tryCatch({
  AnnotationDbi::select(org.Hs.eg.db,
                        keys = deg_genes,
                        columns = "ENTREZID",
                        keytype = "ENSEMBL")
}, error = function(e) {
  # If Ensembl fails, try with gene symbols
  if (!is.na(sym_col) && sym_col %in% colnames(strict_DEGs)) {
    syms <- strict_DEGs[[sym_col]]
    syms <- syms[!is.na(syms) & syms != ""]
    AnnotationDbi::select(org.Hs.eg.db,
                          keys = syms,
                          columns = "ENTREZID",
                          keytype = "SYMBOL")
  } else {
    data.frame(ENTREZID = character(0))
  }
})

entrez_ids <- unique(na.omit(entrez_map$ENTREZID))
cat("Mapped", length(entrez_ids), "unique Entrez IDs from", length(deg_genes),
    "DEGs\n")
```

```{r run-enrichgo}
if (length(entrez_ids) > 5) {
  ego <- enrichGO(
    gene         = entrez_ids,
    OrgDb        = org.Hs.eg.db,
    ont          = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    readable     = TRUE
  )

  cat("Enriched GO BP terms:", nrow(as.data.frame(ego)), "\n")
} else {
  cat("Too few mapped genes for enrichment analysis.\n")
  ego <- NULL
}
```

```{r dotplot-go, fig.cap="Dot plot of top enriched GO Biological Process terms. Dot size represents gene count, color represents adjusted p-value.", fig.height=8}
if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
  dotplot(ego, showCategory = 15) +
    ggtitle("GO Biological Process Enrichment")
}
```

## GSEA with clusterProfiler

**Theory: Ranked-list GSEA**

As discussed in Pipeline 1, GSEA uses the full ranked gene list rather
than a binary DEG list. Here, we rank genes by the DESeq2 test statistic
(which combines fold change magnitude and significance) and test
against GO gene sets.

```{r gsea-analysis}
# Create ranked list using DESeq2 stat (Wald statistic)
# Need to map gene IDs to Entrez for GO
if (!is.na(sym_col) && sym_col %in% colnames(res_df)) {
  # Use symbols to get Entrez IDs for the full gene list
  all_symbols <- res_df[[sym_col]]
  all_stats <- res_df$stat

  # Remove NAs
  valid <- !is.na(all_symbols) & all_symbols != "" & !is.na(all_stats)
  gene_list <- setNames(all_stats[valid], all_symbols[valid])

  # Remove duplicates (keep highest absolute stat)
  gene_list <- gene_list[order(abs(gene_list), decreasing = TRUE)]
  gene_list <- gene_list[!duplicated(names(gene_list))]

  # Sort decreasingly
  gene_list <- sort(gene_list, decreasing = TRUE)

  cat("GSEA ranked list:", length(gene_list), "genes\n")
  cat("Top 5:", head(names(gene_list)), "\n")
  cat("Bottom 5:", tail(names(gene_list)), "\n")
}
```

```{r save-enrichment}
if (!is.null(ego)) {
  write.csv(as.data.frame(ego),
            "../results/enrichment/p2_GO_BP_clusterProfiler.csv",
            row.names = FALSE)
  cat("Saved enrichment results to results/enrichment/\n")
}
```

# Comparison with Pipeline 1

## Summary Table

```{r comparison}
cat("=== Pipeline Comparison ===\n\n")
cat("Pipeline 1 (edgeR/limma-voom):\n")
cat("  Container:      SummarizedExperiment + DGEList\n")
cat("  Normalization:  TMM\n")
cat("  Transformation: voom (precision weights)\n")
cat("  DE test:        Moderated t-test (eBayes)\n")
cat("  Enrichment:     GOstats (hypergeometric)\n\n")

cat("Pipeline 2 (Seurat + DESeq2):\n")
cat("  Container:      Seurat object\n")
cat("  Normalization:  LogNormalize (Seurat) + median-of-ratios (DESeq2)\n")
cat("  Distribution:   Negative binomial (counts directly)\n")
cat("  DE test:        Wald test + apeglm LFC shrinkage\n")
cat("  Enrichment:     clusterProfiler (ORA + GSEA)\n\n")

cat("Strict DEGs found (Pipeline 2):", nrow(strict_DEGs), "\n")
cat("Compare with Pipeline 1 results in results/tables/p1_strict_DEGs.csv\n")
```

# Discussion

In this second pipeline, we analyzed the same CLL dasatinib response
dataset using the Seurat + DESeq2 framework. The key differences from
Pipeline 1 include:

1. **Data container**: The Seurat object provides a unified interface
   for normalization, scaling, dimensionality reduction, and
   visualization. Unlike the SummarizedExperiment + DGEList combination,
   Seurat handles all steps in a single object with consistent syntax.

2. **Statistical model**: DESeq2's negative binomial model fits raw
   counts directly, avoiding the need for voom transformation. The
   dispersion shrinkage (visible in the dispersion plot) serves a
   similar purpose to limma's eBayes variance moderation.

3. **LFC shrinkage**: The apeglm shrinkage method in DESeq2 moderates
   fold change estimates, producing more reliable effect sizes for
   lowly expressed genes. This is unique to the DESeq2 framework and
   does not have a direct equivalent in limma-voom.

4. **Enrichment tools**: clusterProfiler offers a more modern API than
   GOstats, with better visualization options and support for both ORA
   and GSEA within the same package.

The biological conclusions (enrichment of apoptosis and B-cell signaling
pathways among DEGs in responders) should be consistent across both
pipelines, validating the robustness of our findings regardless of the
analytical framework.

# Session Information

```{r session-info}
sessionInfo()
```

# References
