---
title: "FASTQ to Count Matrix: Complete Salmon Pipeline"
subtitle: "From raw paired-end reads to a gene-level count matrix in R"
author: "Samuel Escudero"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

This guide walks through every step of processing 14 paired-end bulk
RNA-seq samples using **Salmon** for pseudo-alignment and
quantification, followed by **tximeta** in R to import transcript-level
abundances and summarize them to gene-level counts. We produce a
`SummarizedExperiment` object saved as `.rds` for reuse in downstream
pipelines (edgeR, DESeq2, Seurat, etc.).

**What we cover:**

1. Compressing and transferring FASTQ files to the cluster
2. Installing Salmon and FastQC on the cluster
3. Downloading the GRCh38 reference transcriptome and genome
4. Building the Salmon index (with decoy sequences)
5. Quality control with FastQC + MultiQC
6. Quantifying all 14 samples with Salmon (loop)
7. Importing into R with tximeta
8. Summarizing to gene level
9. Saving as RDS for downstream analysis

**Approximate total time:** 2-3 hours for 14 samples (depending on
cluster speed and queue times).

# Step 0: Prepare and Transfer FASTQ Files

## On your local computer (BASH)

First, we compress our FASTQ files (if they are not already .gz) and
transfer them to the cluster.

```bash
# ============================================================
# RUN THIS ON YOUR LOCAL MACHINE (laptop/desktop)
# ============================================================

# Check your FASTQ files
# They should be named like: sample1_R1.fastq, sample1_R2.fastq
# or already compressed: sample1_R1.fastq.gz
ls raw_data/
# Expected: 28 files (14 samples x 2 reads each: R1 and R2)

# If your files are NOT compressed (.fastq without .gz), compress them
# This saves ~70% disk space and is required for transfer
# Time: ~2-5 min per file depending on size
cd raw_data/
for f in *.fastq; do
    echo "Compressing ${f}..."
    gzip "${f}"
done

# Verify all files are compressed
ls *.fastq.gz | wc -l
# Should show 28

# Create a tar archive of all compressed FASTQs for faster transfer
# Time: ~1-2 min
cd ..
tar -cf fastq_archive.tar raw_data/*.fastq.gz
echo "Archive created: $(du -h fastq_archive.tar | cut -f1)"
```

## Transfer to the cluster (BASH - local machine)

```bash
# ============================================================
# RUN THIS ON YOUR LOCAL MACHINE
# Transfer the archive to the cluster via scp
# Replace <username> and <cluster_address> with your details
# Time: depends on network speed, typically 10-30 min for ~50 GB
# ============================================================

scp fastq_archive.tar <username>@<cluster_address>:~/projects/rnaseq_project/

# Alternative: if you prefer rsync (can resume interrupted transfers)
# rsync -avP fastq_archive.tar <username>@<cluster_address>:~/projects/rnaseq_project/
```

## On the cluster: unpack the archive (BASH)

```bash
# ============================================================
# RUN THIS ON THE CLUSTER (after ssh login)
# ============================================================

# Connect to the cluster
ssh <username>@<cluster_address>

# Navigate to your project directory
mkdir -p ~/projects/rnaseq_project
cd ~/projects/rnaseq_project

# Create the directory structure
mkdir -p raw_data
mkdir -p reference
mkdir -p salmon_index
mkdir -p salmon_quant
mkdir -p qc_results
mkdir -p results

# Extract the archive
# Time: ~1-2 min
tar -xf fastq_archive.tar
# If files extracted into a subdirectory, move them:
# mv raw_data/*.fastq.gz ./raw_data/

# Verify
ls raw_data/*.fastq.gz | wc -l
# Should show 28 files (14 samples x 2 reads)

# Check the sample names
ls raw_data/*_R1.fastq.gz | sed 's/_R1.fastq.gz//' | sed 's|raw_data/||'
# This prints your 14 sample names
```

# Step 1: Install Tools on the Cluster

## Load or install Salmon and FastQC (BASH - cluster)

Most HPC clusters have these available via module systems. Check first:

```bash
# ============================================================
# RUN ON THE CLUSTER
# ============================================================

# Option A: Load pre-installed modules (most common on HPC)
module avail salmon
module avail fastqc
module avail multiqc

# If available, load them:
module load salmon/1.10.0    # or whatever version is available
module load fastqc/0.12.1
module load multiqc/1.14

# Verify
salmon -v
fastqc -v

# Option B: If not available via modules, install with conda
# First, make sure conda/miniconda is available
module load miniconda3   # or anaconda3

# Create a dedicated environment
conda create -n rnaseq_env -c bioconda -c conda-forge \
    salmon=1.10 fastqc multiqc

conda activate rnaseq_env

# Verify installation
salmon -v
# Expected output: salmon 1.10.x
fastqc -v
# Expected output: FastQC v0.12.x
```

# Step 2: Download the GRCh38 Reference

## Download transcriptome and genome from ENSEMBL (BASH - cluster)

Salmon needs the transcriptome FASTA (cDNA sequences) for
pseudo-alignment. For best results, we also download the genome FASTA
to use as decoy sequences. Decoys prevent spurious mappings of reads
that originate from intergenic or intronic regions.

```bash
# ============================================================
# RUN ON THE CLUSTER
# Time: ~10-20 min depending on download speed
# ============================================================

cd ~/projects/rnaseq_project/reference

# Download the human transcriptome (cDNA) from ENSEMBL release 109
# This is the set of all known mRNA transcript sequences
# Size: ~300 MB compressed
wget -O Homo_sapiens.GRCh38.cdna.all.fa.gz \
    https://ftp.ensembl.org/pub/release-109/fasta/homo_sapiens/cdna/Homo_sapiens.GRCh38.cdna.all.fa.gz

# Download the primary genome assembly (for decoy sequences)
# Size: ~900 MB compressed
wget -O Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz \
    https://ftp.ensembl.org/pub/release-109/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz

# Download the GTF annotation file (needed later for tximeta in R)
# Size: ~50 MB compressed
wget -O Homo_sapiens.GRCh38.109.gtf.gz \
    https://ftp.ensembl.org/pub/release-109/gtf/homo_sapiens/Homo_sapiens.GRCh38.109.gtf.gz

# Decompress the GTF (we need it uncompressed later)
gunzip Homo_sapiens.GRCh38.109.gtf.gz

# Verify downloads
ls -lh reference/
# Expected:
# Homo_sapiens.GRCh38.cdna.all.fa.gz        (~300 MB)
# Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz (~900 MB)
# Homo_sapiens.GRCh38.109.gtf               (~1.4 GB uncompressed)
```

## Prepare decoy sequences (BASH - cluster)

Salmon's decoy-aware mode improves quantification accuracy. We create
a "gentrome" file (transcriptome + genome concatenated) and a list of
decoy chromosome names.

```bash
# ============================================================
# RUN ON THE CLUSTER
# Time: ~5-10 min
# ============================================================

cd ~/projects/rnaseq_project/reference

# Extract chromosome names from the genome as decoy names
# These tell Salmon which sequences are decoys (genomic, not transcripts)
grep "^>" <(gunzip -c Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz) \
    | cut -d " " -f 1 | sed 's/>//' > decoys.txt

# Check: should list chromosome names (1, 2, 3, ..., X, Y, MT, etc.)
head decoys.txt
wc -l decoys.txt
# Expected: ~190-640 entries (chromosomes + scaffolds)

# Create the gentrome: concatenate transcriptome + genome
# IMPORTANT: transcriptome MUST come first
cat Homo_sapiens.GRCh38.cdna.all.fa.gz \
    Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz \
    > gentrome.fa.gz

echo "Gentrome created: $(du -h gentrome.fa.gz | cut -f1)"
```

# Step 3: Build the Salmon Index

```bash
# ============================================================
# RUN ON THE CLUSTER
# Time: ~20-40 min, requires ~16 GB RAM
# Submit as a job if your cluster requires it (see SLURM example below)
# ============================================================

cd ~/projects/rnaseq_project

# Build the decoy-aware Salmon index
# -t: input FASTA (gentrome = transcriptome + genome)
# -d: decoy sequences file
# -i: output index directory
# -p: number of threads
salmon index \
    -t reference/gentrome.fa.gz \
    -d reference/decoys.txt \
    -i salmon_index \
    -p 8

# Verify the index was created
ls salmon_index/
# Expected: several files including info.json, pos.bin, etc.

echo "Salmon index complete."
```

**SLURM job script** (if your cluster uses SLURM):

```bash
#!/bin/bash
#SBATCH -J salmon_index
#SBATCH -o salmon_index_%j.out
#SBATCH -e salmon_index_%j.err
#SBATCH -p normal
#SBATCH -N 1
#SBATCH -n 8
#SBATCH -t 02:00:00
#SBATCH -A your_allocation

module load salmon/1.10.0

cd ~/projects/rnaseq_project

salmon index \
    -t reference/gentrome.fa.gz \
    -d reference/decoys.txt \
    -i salmon_index \
    -p 8
```

Save this as `scripts/01_salmon_index.sh` and submit with `sbatch scripts/01_salmon_index.sh`.

# Step 4: Quality Control with FastQC + MultiQC

```bash
# ============================================================
# RUN ON THE CLUSTER
# Time: ~15-30 min for 28 files (14 samples x 2 reads)
# ============================================================

cd ~/projects/rnaseq_project

# Run FastQC on all FASTQ files
# -o: output directory
# -t: number of threads
fastqc -o qc_results/ -t 8 raw_data/*.fastq.gz

# Aggregate all FastQC reports into a single MultiQC report
multiqc qc_results/ -o qc_results/multiqc_report

# The HTML report is at:
# qc_results/multiqc_report/multiqc_report.html
# Download this to your local machine to view in a browser:
# scp <user>@<cluster>:~/projects/rnaseq_project/qc_results/multiqc_report/multiqc_report.html .

echo "QC complete. Review the MultiQC report before proceeding."
```

**What to look for in the MultiQC report:**

- Per-base sequence quality: should be >28 across all positions
- Adapter content: should be minimal (<5%). If high, trim with fastp
- GC content: should follow a normal distribution
- Sequence duplication: moderate duplication is normal for RNA-seq
- Overrepresented sequences: check for adapter contamination

If you need to trim adapters (only if MultiQC shows problems):

```bash
# Optional: trim adapters with fastp (only if needed)
# Time: ~2-3 min per sample
for sample in $(ls raw_data/*_R1.fastq.gz | sed 's/_R1.fastq.gz//' | sed 's|raw_data/||'); do
    fastp \
        -i raw_data/${sample}_R1.fastq.gz \
        -I raw_data/${sample}_R2.fastq.gz \
        -o trimmed/${sample}_R1_trimmed.fastq.gz \
        -O trimmed/${sample}_R2_trimmed.fastq.gz \
        -h qc_results/${sample}_fastp.html \
        -w 4
done
# If trimming was done, use trimmed/ files for salmon quant below
```

# Step 5: Quantify All 14 Samples with Salmon

This is the core step. Salmon performs pseudo-alignment and
quantification in a single pass, producing per-sample directories with
transcript-level counts, TPM values, and effective lengths.

```bash
# ============================================================
# RUN ON THE CLUSTER
# Time: ~3-8 min per sample, ~45-120 min total for 14 samples
# ============================================================

cd ~/projects/rnaseq_project

# Create a list of sample names from the R1 files
# Adjust the pattern to match YOUR file naming convention
SAMPLES=$(ls raw_data/*_R1.fastq.gz | sed 's/_R1.fastq.gz//' | sed 's|raw_data/||')

echo "Samples to process:"
echo "${SAMPLES}"
echo "Total: $(echo "${SAMPLES}" | wc -w) samples"

# Loop through all samples and quantify
for sample in ${SAMPLES}; do
    echo "==========================================="
    echo "Processing: ${sample}"
    echo "Start time: $(date)"
    echo "==========================================="

    salmon quant \
        -i salmon_index \
        -l A \
        -1 raw_data/${sample}_R1.fastq.gz \
        -2 raw_data/${sample}_R2.fastq.gz \
        -p 8 \
        -o salmon_quant/${sample} \
        -g reference/Homo_sapiens.GRCh38.109.gtf \
        -validateMappings

    echo "Finished: ${sample} at $(date)"
    echo ""
done

echo "All samples quantified."
```

**SLURM array job** (recommended for parallel processing):

```bash
#!/bin/bash
#SBATCH -J salmon_quant
#SBATCH -o salmon_quant_%A_%a.out
#SBATCH -e salmon_quant_%A_%a.err
#SBATCH -p normal
#SBATCH -N 1
#SBATCH -n 8
#SBATCH -t 01:00:00
#SBATCH -A your_allocation
#SBATCH -a 1-14

module load salmon/1.10.0

cd ~/projects/rnaseq_project

# Get the sample name for this array task
SAMPLE=$(ls raw_data/*_R1.fastq.gz | sed 's/_R1.fastq.gz//' | sed 's|raw_data/||' | sed -n "${SLURM_ARRAY_TASK_ID}p")

echo "Processing sample ${SLURM_ARRAY_TASK_ID}: ${SAMPLE}"

salmon quant \
    -i salmon_index \
    -l A \
    -1 raw_data/${SAMPLE}_R1.fastq.gz \
    -2 raw_data/${SAMPLE}_R2.fastq.gz \
    -p 8 \
    -o salmon_quant/${SAMPLE} \
    -g reference/Homo_sapiens.GRCh38.109.gtf \
    -validateMappings
```

Save as `scripts/02_salmon_quant_array.sh` and submit:
`sbatch scripts/02_salmon_quant_array.sh`
This runs all 14 samples in parallel (total time: ~8 min if slots are
available).

## Verify Salmon output

```bash
# ============================================================
# RUN ON THE CLUSTER
# ============================================================

# Check that all 14 samples have quant.sf files
ls salmon_quant/*/quant.sf | wc -l
# Expected: 14

# Check mapping rates
for d in salmon_quant/*/; do
    sample=$(basename ${d})
    rate=$(grep "Mapping rate" ${d}/aux_info/meta_info.json 2>/dev/null || echo "N/A")
    echo "${sample}: ${rate}"
done

# Quick look at a quant.sf file
head salmon_quant/$(ls salmon_quant/ | head -1)/quant.sf
# Columns: Name, Length, EffectiveLength, TPM, NumReads
```

# Step 6: Import into R with tximeta

**Now we switch to R.** Everything below runs in R (RStudio or
an R script on the cluster).

## Create the sample table

```{r import-setup}
# ============================================================
# RUN IN R (RStudio or R on the cluster)
# ============================================================

# Install packages if needed (run once)
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("tximeta", "SummarizedExperiment", "DESeq2"))

library(tximeta)
library(SummarizedExperiment)

# Define the project directory
project_dir <- "~/projects/rnaseq_project"

# Build the sample table
# List all quant.sf files
quant_files <- list.files(
  file.path(project_dir, "salmon_quant"),
  pattern = "quant.sf",
  recursive = TRUE,
  full.names = TRUE
)

# Extract sample names from directory names
sample_names <- basename(dirname(quant_files))

cat("Found", length(quant_files), "quantification files:\n")
print(sample_names)

# Create the coldata data frame that tximeta requires
# IMPORTANT: adjust the "condition" column to match YOUR experiment
# Here we show a generic example with two groups
coldata <- data.frame(
  names     = sample_names,
  files     = quant_files,
  condition = factor(c(
    # REPLACE THIS with your actual group assignments
    # Example for 14 samples: 7 control + 7 treatment
    rep("Control", 7),
    rep("Treatment", 7)
  )),
  stringsAsFactors = FALSE
)

# Verify
print(coldata)
```

## Import with tximeta

**Theory: What does tximeta do?**

`tximeta` is an enhanced version of `tximport` that automatically
detects which reference transcriptome was used for quantification
(by checking Salmon's metadata) and links it to the correct gene
annotations. This means we get gene IDs, gene names, and genomic
coordinates automatically, without having to manually provide a
tx2gene mapping file.

If tximeta cannot auto-detect the reference, we can provide the
annotation manually using `makeLinkedTxome()`.

```{r run-tximeta}
# ============================================================
# RUN IN R
# Time: ~1-3 min
# ============================================================

# Import transcript-level quantifications
# tximeta auto-detects the reference and downloads annotation
se <- tximeta(coldata)

cat("Transcript-level SummarizedExperiment created:\n")
se
```

```{r manual-txome, eval=FALSE}
# If tximeta auto-detection fails, create a linked transcriptome manually:
# This tells tximeta exactly which reference was used

gtf_path <- file.path(project_dir, "reference", "Homo_sapiens.GRCh38.109.gtf")
fasta_path <- file.path(project_dir, "reference", "Homo_sapiens.GRCh38.cdna.all.fa.gz")
index_dir <- file.path(project_dir, "salmon_index")

makeLinkedTxome(
  indexDir   = index_dir,
  source     = "Ensembl",
  organism   = "Homo sapiens",
  release    = "109",
  genome     = "GRCh38",
  fasta      = fasta_path,
  gtf        = gtf_path,
  write      = FALSE
)

# Then re-run tximeta
se <- tximeta(coldata)
```

## Summarize to gene level

```{r summarize-genes}
# ============================================================
# RUN IN R
# Time: ~30 sec
# ============================================================

# Summarize transcript-level counts to gene-level counts
# This sums all transcript counts belonging to the same gene
gse <- summarizeToGene(se)

cat("Gene-level SummarizedExperiment:\n")
gse

cat("\nDimensions:", nrow(gse), "genes x", ncol(gse), "samples\n")
cat("\nAssays available:", names(assays(gse)), "\n")
cat("  - counts: raw counts (for DESeq2/edgeR)\n")
cat("  - abundance: TPM values\n")
cat("  - length: effective gene lengths\n")

cat("\nSample metadata:\n")
print(colData(gse))

cat("\nFirst 5 genes:\n")
head(rowData(gse), 5)
```

## Inspect the count matrix

```{r inspect-counts}
# ============================================================
# RUN IN R
# ============================================================

# Look at the count matrix
counts_matrix <- assay(gse, "counts")
cat("Count matrix dimensions:", dim(counts_matrix), "\n")
cat("Total genes:", nrow(counts_matrix), "\n")
cat("Total samples:", ncol(counts_matrix), "\n\n")

# Library sizes (total counts per sample)
lib_sizes <- colSums(counts_matrix)
cat("Library sizes (millions of reads):\n")
print(round(lib_sizes / 1e6, 2))

# Detected genes per sample (count > 0)
detected <- colSums(counts_matrix > 0)
cat("\nDetected genes per sample:\n")
print(detected)

# Quick summary of the count distribution
cat("\nCount matrix summary (first 3 samples):\n")
print(summary(counts_matrix[, 1:3]))
```

# Step 7: Save as RDS

## Save the SummarizedExperiment object

```{r save-rds}
# ============================================================
# RUN IN R
# Time: ~10-30 sec
# ============================================================

# Save the gene-level SummarizedExperiment as an RDS file
# This preserves the entire object: counts, metadata, gene annotations
output_path <- file.path(project_dir, "results", "gene_counts_se.rds")
saveRDS(gse, file = output_path)

cat("SummarizedExperiment saved to:", output_path, "\n")
cat("File size:", round(file.size(output_path) / 1e6, 1), "MB\n")

# Also save just the count matrix as CSV (for use in Python/other tools)
write.csv(
  as.data.frame(counts_matrix),
  file.path(project_dir, "results", "gene_counts_matrix.csv"),
  row.names = TRUE
)

# Save sample metadata
write.csv(
  as.data.frame(colData(gse)),
  file.path(project_dir, "results", "sample_metadata.csv"),
  row.names = TRUE
)

cat("CSV files saved to results/\n")
```

## How to load the RDS later

```{r load-rds}
# ============================================================
# RUN IN R (every time you start a new analysis session)
# ============================================================

library(SummarizedExperiment)

# Load the saved object
gse <- readRDS("~/projects/rnaseq_project/results/gene_counts_se.rds")

# Verify it loaded correctly
cat("Loaded:", class(gse), "\n")
cat("Dimensions:", nrow(gse), "genes x", ncol(gse), "samples\n")
cat("Assays:", names(assays(gse)), "\n")

# Access the count matrix
counts_matrix <- assay(gse, "counts")

# Access sample metadata
sample_info <- colData(gse)

# Access gene annotations
gene_info <- rowData(gse)

# Now you can proceed with DESeq2, edgeR, or Seurat
# Example: create a DESeqDataSet
library(DESeq2)
dds <- DESeqDataSet(gse, design = ~ condition)
```

# Step 8: Quick Sanity Check

```{r sanity-check}
# ============================================================
# RUN IN R
# Quick checks before downstream analysis
# ============================================================

library(ggplot2)

# 1. Library size barplot
lib_df <- data.frame(
  sample = colnames(gse),
  lib_size = colSums(assay(gse, "counts")) / 1e6,
  condition = colData(gse)$condition
)

ggplot(lib_df, aes(x = reorder(sample, lib_size), y = lib_size, fill = condition)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Sample", y = "Library size (millions)", fill = "Condition") +
  theme_minimal()

# 2. Gene detection rate
cat("\nGenes with > 0 counts in all samples:",
    sum(rowSums(assay(gse, "counts") > 0) == ncol(gse)), "\n")
cat("Genes with > 10 counts total:",
    sum(rowSums(assay(gse, "counts")) > 10), "\n")
```

# Summary of Steps and Times

| Step | Tool | Where | Approx. Time |
|------|------|-------|-------------|
| 0. Compress FASTQs | gzip | Local | 2-5 min/file |
| 0. Transfer to cluster | scp/rsync | Local | 10-30 min |
| 0. Unpack archive | tar | Cluster (bash) | 1-2 min |
| 1. Install tools | module/conda | Cluster (bash) | 5 min |
| 2. Download references | wget | Cluster (bash) | 10-20 min |
| 2. Prepare decoys | grep/cat | Cluster (bash) | 5-10 min |
| 3. Build Salmon index | salmon index | Cluster (bash) | 20-40 min |
| 4. FastQC + MultiQC | fastqc/multiqc | Cluster (bash) | 15-30 min |
| 5. Salmon quant (x14) | salmon quant | Cluster (bash) | 3-8 min/sample |
| 6. Import to R | tximeta | R (cluster or local) | 1-3 min |
| 6. Summarize to gene | summarizeToGene | R | 30 sec |
| 7. Save RDS | saveRDS | R | 10-30 sec |

**Total: approximately 2-3 hours** (or ~30 min if Salmon quant is
parallelized with SLURM array jobs).

# References

- Patro R et al. (2017). Salmon provides fast and bias-aware
  quantification of transcript expression. Nature Methods 14:417-419.
- Love MI et al. (2020). tximeta: Reference sequence checksums for
  provenance identification in RNA-seq. PLOS Computational Biology.
- Soneson C et al. (2015). Differential analyses for RNA-seq:
  transcript-level estimates improve gene-level inferences. F1000Research.

# Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
